---
title: "Experiment 1: Cost Pressure"
author: "Elves Project Team"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 8
    fig-height: 5
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(tidyboot)
library(lme4)
library(lmerTest)
library(here)
library(patchwork)

# Color palettes
colors_tp <- c("5" = "#E94F37", "10" = "#2E86AB")
colors_freq <- c("1:2" = "#6B4C9A", "1:4" = "#2E86AB")

# Output paths
fig_path <- here("latex/figures")
```

# Data Preprocessing

```{r load-data}
# Load all raw data once, selecting only columns we need to avoid type conflicts
raw_data_path <- here("data/full-production/raw_data/exp-1")
raw_files <- list.files(raw_data_path, full.names = TRUE) |>
  str_subset("inprog", negate = TRUE)

cols_needed <- c("subject_id", "frequency_ratio", "time_pressure", "survey_difficulty",
                 "is_correct", "response", "response_text", "target", "angle", "targetFreq",
                 "rt", "nearest_trained_angle", "timed_out")

raw_data <- map_dfr(raw_files, \(f) {
  read_csv(f, show_col_types = FALSE) |>
    select(any_of(cols_needed)) |>
    mutate(across(everything(), as.character))  # coerce all to character, convert later
})

cat("Loaded", n_distinct(raw_data$subject_id), "participants\n")
```

```{r extract-data}
# Extract participant conditions
participants <- raw_data |>
  filter(!is.na(survey_difficulty)) |>
  distinct(subject_id, frequency_ratio, time_pressure) |>
  mutate(time_pressure = as.numeric(time_pressure))

# Extract angle-word mappings from exposure phase
angle_mappings <- raw_data |>
  filter(!is.na(is_correct), str_detect(response, "exposure-response")) |>
  filter(is_correct == "1") |>
  distinct(subject_id, angle = as.numeric(angle), target, targetFreq)

# Extract production trials
production <- raw_data |>
  filter(!is.na(nearest_trained_angle)) |>
  transmute(
    subject_id,
    rt = as.numeric(rt),
    response_text, target,
    angle = as.numeric(angle),
    targetFreq,
    nearest_trained_angle = as.numeric(nearest_trained_angle),
    timed_out = as.numeric(timed_out)
  ) |>
  mutate(
    # Circular distance to nearest (handles 0°/360° wraparound)
    distance = pmin(abs(angle - nearest_trained_angle), 360 - abs(angle - nearest_trained_angle)),
    # Compute both potential neighbors (±45° from nearest)
    neighbor_plus = (nearest_trained_angle + 45) %% 360,
    neighbor_minus = (nearest_trained_angle - 45 + 360) %% 360,
    # Circular distance to each neighbor
    dist_plus = pmin(abs(angle - neighbor_plus), 360 - abs(angle - neighbor_plus)),
    dist_minus = pmin(abs(angle - neighbor_minus), 360 - abs(angle - neighbor_minus)),
    # Next-nearest is the closer neighbor (handles 0°/360° wraparound)
    next_nearest_angle = ifelse(dist_plus < dist_minus, neighbor_plus, neighbor_minus)
  ) |>
  select(-neighbor_plus, -neighbor_minus, -dist_plus, -dist_minus)
```

```{r join-frequencies}
# Build lookup table for angle -> frequency mapping
angle_freq_lookup <- angle_mappings |>
  select(subject_id, angle, freq = targetFreq, label = target)

# Join frequency labels for nearest, next-nearest, and response
d_exp1 <- production |>
  left_join(angle_freq_lookup |> rename(nearestAngle = angle, nearestFreq = freq, nearestLabel = label),
            by = c("subject_id", "nearest_trained_angle" = "nearestAngle")) |>
  left_join(angle_freq_lookup |> rename(nextNearestFreq = freq, nextNearestLabel = label),
            by = c("subject_id", "next_nearest_angle" = "angle")) |>
  left_join(angle_freq_lookup |> select(subject_id, response_text = label, responseFreq = freq),
            by = c("subject_id", "response_text")) |>
  left_join(participants, by = "subject_id") |>
  mutate(
    between_freqs = nearestFreq != nextNearestFreq,
    # Ambiguous trial: test angle is past midpoint (>11° from nearest, toward next-nearest)
    ambig_trial = distance > 11,
    # Critical trial: between HF/LF frequencies AND in ambiguous zone
    is_critical = between_freqs & ambig_trial,
    # Distance to LF word (key predictor)
    # distance is already circular, and next_nearest uses circular distance
    dist_to_lf = ifelse(
      nearestFreq == "LF",
      distance,
      pmin(abs(angle - next_nearest_angle), 360 - abs(angle - next_nearest_angle))
    ),
    # LF word label for this trial (for random effects)
    lf_word = ifelse(nearestFreq == "LF", nearestLabel, nextNearestLabel)
  )

cat("Total trials:", nrow(d_exp1), "\n")
cat("Between-frequency trials:", sum(d_exp1$between_freqs, na.rm = TRUE), "\n")
cat("Ambiguous trials:", sum(d_exp1$ambig_trial, na.rm = TRUE), "\n")
cat("Critical trials:", sum(d_exp1$is_critical, na.rm = TRUE), "\n")
```

# Sample Descriptives

```{r sample-descriptives}
n_participants <- n_distinct(participants$subject_id)

participants |>
  count(time_pressure, frequency_ratio) |>
  pivot_wider(names_from = frequency_ratio, values_from = n) |>
  knitr::kable(col.names = c("Time Pressure (s)", "1:2", "1:4"),
               caption = sprintf("Participants per condition (N = %d)", n_participants))
```

```{r timeout-rates}
timeout_by_tp_exp1 <- d_exp1 |>
  group_by(time_pressure) |>
  summarise(
    n_trials = n(),
    n_timeout = sum(timed_out, na.rm = TRUE),
    pct_timeout = 100 * n_timeout / n_trials,
    .groups = "drop"
  )

knitr::kable(timeout_by_tp_exp1, digits = 1,
             col.names = c("Time Pressure (s)", "N Trials", "N Timeouts", "% Timeout"))
```

# Critical Trial Selection

```{r filter-critical}
d_critical_exp1 <- d_exp1 |>
  filter(between_freqs, timed_out == 0, !is.na(responseFreq)) |>
  mutate(
    responded_lf = as.numeric(responseFreq == "LF"),
    time_pressure_c = ifelse(time_pressure == 5, -0.5, 0.5),
    freq_ratio_c = ifelse(frequency_ratio == "1:2", -0.5, 0.5),
    # Scale distance for model stability
    dist_to_lf_scaled = scale(dist_to_lf)[,1]
  )

cat("Critical trials:", nrow(d_critical_exp1), "\n")
cat("N participants:", n_distinct(d_critical_exp1$subject_id), "\n")
cat("N unique LF words:", n_distinct(d_critical_exp1$lf_word), "\n")
cat("Overall P(LF):", round(mean(d_critical_exp1$responded_lf), 3), "\n")
```

```{r}
hist(d_critical_exp1$dist_to_lf_scaled)
```


# Statistical Models

```{r models}
# Time pressure × distance interaction
# Random effects: by-participant intercepts and slopes for distance (uncorrelated), by-word intercepts
m_tp_exp1 <- glmer(
  responded_lf ~ time_pressure_c * dist_to_lf_scaled + freq_ratio_c + 
    (1 + dist_to_lf_scaled | subject_id) + (1 | lf_word),
  data = d_critical_exp1, family = binomial,
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

# Frequency ratio × distance interaction
m_freq_exp1 <- glmer(
  responded_lf ~ freq_ratio_c * dist_to_lf_scaled + time_pressure_c + 
    (1 + dist_to_lf_scaled | subject_id) + (1 | lf_word),
  data = d_critical_exp1, family = binomial,
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

tp_coef_exp1 <- summary(m_tp_exp1)$coefficients["time_pressure_c:dist_to_lf_scaled", ]
freq_coef_exp1 <- summary(m_freq_exp1)$coefficients["freq_ratio_c:dist_to_lf_scaled", ]
```

```{r model-summaries}
summary(m_tp_exp1)
summary(m_freq_exp1)
```

# Model: Chose Nearest by Target Frequency

This analysis examines whether participants chose the nearest trained angle as a function of whether that angle was HF or LF.

```{r model-chose-nearest-exp1}
# Prepare data for chose-nearest analysis
d_chose_nearest_exp1 <- d_exp1 |>
  filter(between_freqs, timed_out == 0, !is.na(responseFreq)) |>
  mutate(
    # Outcome: did participant respond with the nearest trained angle's word?
    chose_nearest = as.numeric(response_text == nearestLabel),
    # Predictor: is the nearest angle HF? (centered: -0.5 = LF, 0.5 = HF)
    is_hf_trial_c = ifelse(nearestFreq == "HF", 0.5, -0.5),
    # Centered predictors
    time_pressure_c = ifelse(time_pressure == 5, -0.5, 0.5),
    freq_ratio_c = ifelse(frequency_ratio == "1:2", -0.5, 0.5),
    distance_c = scale(distance)[,1]
  )

cat("Chose-nearest analysis:\n")
cat("  N trials:", nrow(d_chose_nearest_exp1), "\n")
cat("  N participants:", n_distinct(d_chose_nearest_exp1$subject_id), "\n")
cat("  Overall P(chose nearest):", round(mean(d_chose_nearest_exp1$chose_nearest), 3), "\n")
cat("  P(chose nearest | HF nearest):", round(mean(d_chose_nearest_exp1$chose_nearest[d_chose_nearest_exp1$nearestFreq == "HF"]), 3), "\n")
cat("  P(chose nearest | LF nearest):", round(mean(d_chose_nearest_exp1$chose_nearest[d_chose_nearest_exp1$nearestFreq == "LF"]), 3), "\n")
```

```{r model-chose-nearest-fit-exp1}
# Full model: is_hf_trial × freq_ratio × time_pressure interaction + distance
m_chose_nearest_exp1 <- glmer(
  chose_nearest ~ is_hf_trial_c * freq_ratio_c * time_pressure_c + distance_c +
    (1 + distance_c | subject_id) + (1 | nearestLabel),
  data = d_chose_nearest_exp1, family = binomial,
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

summary(m_chose_nearest_exp1)
```

```{r chose-nearest-by-distance-exp1, fig.width=5, fig.height=4}
# Plot P(chose nearest) by distance, split by HF/LF nearest
n_bins_cn <- 9  # Define here for use in binning

# Filter to 0-22.5 range and bin
d_binned_chose_nearest_exp1 <- d_chose_nearest_exp1 |>
  filter(distance <= 22.5) |>
  mutate(dist_bin = cut(distance, breaks = n_bins_cn, include.lowest = TRUE)) |>
  group_by(nearestFreq, time_pressure, dist_bin) |>
  tidyboot_mean(chose_nearest) |>
  left_join(
    d_chose_nearest_exp1 |>
      filter(distance <= 22.5) |>
      mutate(dist_bin = cut(distance, breaks = n_bins_cn, include.lowest = TRUE)) |>
      group_by(nearestFreq, time_pressure, dist_bin) |>
      summarise(dist_mean = mean(distance), .groups = "drop"),
    by = c("nearestFreq", "time_pressure", "dist_bin")
  )

# Model predictions for the curve
# Get scaling parameters from the data used to fit the model
dist_mean_cn_exp1 <- mean(d_chose_nearest_exp1$distance)
dist_sd_cn_exp1 <- sd(d_chose_nearest_exp1$distance)

pred_chose_nearest_exp1 <- expand.grid(
  distance = seq(0, 22.5, length.out = 100),
  is_hf_trial_c = c(-0.5, 0.5),
  time_pressure_c = c(-0.5, 0.5),
  freq_ratio_c = 0  # Average across frequency ratio conditions
) |>
  mutate(
    distance_c = (distance - dist_mean_cn_exp1) / dist_sd_cn_exp1,
    nearestFreq = ifelse(is_hf_trial_c > 0, "HF", "LF"),
    time_pressure = ifelse(time_pressure_c < 0, 5, 10),
    p_chose_nearest = predict(m_chose_nearest_exp1, newdata = pick(everything()), 
                               type = "response", re.form = NA)
  )

ggplot(d_binned_chose_nearest_exp1, aes(x = dist_mean, y = empirical_stat, 
                                    color = nearestFreq, shape = factor(time_pressure))) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.5) +
  geom_line(data = pred_chose_nearest_exp1, 
            aes(x = distance, y = p_chose_nearest, linetype = factor(time_pressure)),
            linewidth = 1) +
  scale_color_manual(values = c("HF" = "#E94F37", "LF" = "#2E86AB")) +
  scale_linetype_manual(values = c("5" = "solid", "10" = "dashed"),
                        labels = c("5s (Strict)", "10s (Lenient)")) +
  scale_shape_manual(values = c("5" = 16, "10" = 17),
                     labels = c("5s (Strict)", "10s (Lenient)")) +
  scale_x_continuous(limits = c(0, 22.5), breaks = seq(0, 22.5, by = 5)) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.25)) +
  labs(
    x = "Distance from Nearest Trained Angle (deg)",
    y = "P(Chose Nearest)",
    color = "Nearest Freq",
    shape = "Time Pressure",
    linetype = "Time Pressure",
    title = "Exp 1: P(Chose Nearest) by Distance and Target Frequency (Time Pressure)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "right",
    legend.background = element_rect(fill = "white", color = "gray80")
  )
```

```{r chose-nearest-by-distance-freqratio-exp1, fig.width=5, fig.height=4}
# Plot P(chose nearest) by distance, split by HF/LF nearest and FREQUENCY RATIO
n_bins_fr <- 9

# Filter to 0-22.5 range and bin by frequency ratio
d_binned_chose_nearest_fr <- d_chose_nearest_exp1 |>
  filter(distance <= 22.5) |>
  mutate(dist_bin = cut(distance, breaks = n_bins_fr, include.lowest = TRUE)) |>
  group_by(nearestFreq, frequency_ratio, dist_bin) |>
  tidyboot_mean(chose_nearest) |>
  left_join(
    d_chose_nearest_exp1 |>
      filter(distance <= 22.5) |>
      mutate(dist_bin = cut(distance, breaks = n_bins_fr, include.lowest = TRUE)) |>
      group_by(nearestFreq, frequency_ratio, dist_bin) |>
      summarise(dist_mean = mean(distance), .groups = "drop"),
    by = c("nearestFreq", "frequency_ratio", "dist_bin")
  )

# Model predictions for frequency ratio effect
pred_chose_nearest_fr <- expand.grid(
  distance = seq(0, 22.5, length.out = 100),
  is_hf_trial_c = c(-0.5, 0.5),
  freq_ratio_c = c(-0.5, 0.5),
  time_pressure_c = 0  # Average across time pressure conditions
) |>
  mutate(
    distance_c = (distance - dist_mean_cn_exp1) / dist_sd_cn_exp1,
    nearestFreq = ifelse(is_hf_trial_c > 0, "HF", "LF"),
    frequency_ratio = ifelse(freq_ratio_c < 0, "1:2", "1:4"),
    p_chose_nearest = predict(m_chose_nearest_exp1, newdata = pick(everything()), 
                               type = "response", re.form = NA)
  )

ggplot(d_binned_chose_nearest_fr, aes(x = dist_mean, y = empirical_stat, 
                                    color = nearestFreq, shape = frequency_ratio)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.5) +
  geom_line(data = pred_chose_nearest_fr, 
            aes(x = distance, y = p_chose_nearest, linetype = frequency_ratio),
            linewidth = 1) +
  scale_color_manual(values = c("HF" = "#E94F37", "LF" = "#2E86AB")) +
  scale_linetype_manual(values = c("1:2" = "solid", "1:4" = "dashed"),
                        labels = c("1:2", "1:4")) +
  scale_shape_manual(values = c("1:2" = 16, "1:4" = 17),
                     labels = c("1:2", "1:4")) +
  scale_x_continuous(limits = c(0, 22.5), breaks = seq(0, 22.5, by = 5)) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.25)) +
  labs(
    x = "Distance from Nearest Trained Angle (deg)",
    y = "P(Chose Nearest)",
    color = "Nearest Freq",
    shape = "Frequency Ratio",
    linetype = "Frequency Ratio",
    title = "Exp 1: P(Chose Nearest) by Distance and Target Frequency (Freq Ratio)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "right",
    legend.background = element_rect(fill = "white", color = "gray80")
  )
```

# Figure 2: Empirical Results with Race Model Fit

```{r fig-setup}
# Bin data for plotting
n_bins <- 9

# Common plot theme
theme_fig <- theme_minimal(base_size = 12) +
  theme(
    legend.position = c(0.75, 0.85),
    legend.background = element_rect(fill = "white", color = NA),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )

# Race model prediction function (using fitted parameters from models/fit.py)
# Fitted params with separate T for each condition to capture freq × time interaction
# Key finding: ΔT (time pressure effect) is larger for 1:2 than 1:4 (VOC prediction)
race_model_pred <- function(dist_to_lf, freq_ratio, time_pressure) {
  # Convert distance to theta (model space: 0.5 = LF prototype)
  theta <- 0.5 - dist_to_lf / 90

  # Interaction model parameters (8 params, trial-level NLL fit)
  sigma <- 0.194
  delta <- 0.620
  lapse <- 0.157

  # B_HF varies by frequency ratio
  B_hf <- ifelse(freq_ratio == "1:2", 0.497, 0.912)

  # Interaction model: T_5, T_10, gamma
  # T(5s, 1:2) = T_5, T(5s, 1:4) = T_5 + γ, T(10s, *) = T_10
  T_5 <- 3.47
  T_10 <- 5.66
  gamma <- 1.53

  T <- case_when(
    freq_ratio == "1:2" & time_pressure == 5 ~ T_5,
    freq_ratio == "1:2" & time_pressure == 10 ~ T_10,
    freq_ratio == "1:4" & time_pressure == 5 ~ T_5 + gamma,
    freq_ratio == "1:4" & time_pressure == 10 ~ T_10
  )

  # Word locations (HF1=0, LF=0.5, HF2=1)
  word_locs <- c(0, 0.5, 1)

  # Semantic match (Gaussian)
  M <- dnorm(theta, mean = word_locs, sd = sigma)

  # Baseline activations (B_lf = 0 fixed)
  B <- c(B_hf, 0, B_hf)

  # Activation at time T
  A <- B + delta * M * T

  # Softmax + lapse
  probs <- exp(A) / sum(exp(A))
  p_lf <- (1 - lapse) * probs[2] + lapse * (1/3)

  return(p_lf)
}
```

```{r fig-2x2, fig.width=10, fig.height=4.5}
# 2x2 Figure: Split by frequency ratio, time pressure as curves within each panel

# Bin data by all 4 conditions
d_binned_2x2 <- d_critical_exp1 |>
  mutate(dist_bin = cut(dist_to_lf, breaks = n_bins, include.lowest = TRUE)) |>
  group_by(frequency_ratio, time_pressure, dist_bin) |>
  tidyboot_mean(responded_lf) |>
  left_join(
    d_critical_exp1 |>
      mutate(dist_bin = cut(dist_to_lf, breaks = n_bins, include.lowest = TRUE)) |>
      group_by(frequency_ratio, time_pressure, dist_bin) |>
      summarise(dist_mean = mean(dist_to_lf), .groups = "drop"),
    by = c("frequency_ratio", "time_pressure", "dist_bin")
  )

# Race model predictions for smooth curves
pred_race <- expand.grid(
  dist_to_lf = seq(0.5, 44.5, length.out = 100),
  frequency_ratio = c("1:2", "1:4"),
  time_pressure = c(5, 10)
) |>
  rowwise() |>
  mutate(p_lf = race_model_pred(dist_to_lf, frequency_ratio, time_pressure)) |>
  ungroup()

# Panel A: Frequency Ratio 1:2
p_12 <- ggplot(d_binned_2x2 |> filter(frequency_ratio == "1:2"),
               aes(x = dist_mean, y = empirical_stat, color = factor(time_pressure))) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 1) +
  geom_line(data = pred_race |> filter(frequency_ratio == "1:2"),
            aes(x = dist_to_lf, y = p_lf), linewidth = 1.2) +
  geom_vline(xintercept = 22.5, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = colors_tp, labels = c("Strict (5s)", "Lenient (10s)")) +
  scale_x_continuous(limits = c(0, 45), breaks = seq(0, 45, by = 10)) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.25)) +
  labs(x = "Distance from LF (deg)", y = "P(LF)", color = "Time Pressure",
       title = "Frequency Ratio 1:2") +
  theme_fig

# Panel B: Frequency Ratio 1:4
p_14 <- ggplot(d_binned_2x2 |> filter(frequency_ratio == "1:4"),
               aes(x = dist_mean, y = empirical_stat, color = factor(time_pressure))) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 1) +
  geom_line(data = pred_race |> filter(frequency_ratio == "1:4"),
            aes(x = dist_to_lf, y = p_lf), linewidth = 1.2) +
  geom_vline(xintercept = 22.5, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = colors_tp, labels = c("Strict (5s)", "Lenient (10s)")) +
  scale_x_continuous(limits = c(0, 45), breaks = seq(0, 45, by = 10)) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.25)) +
  labs(x = "Distance from LF (deg)", y = "P(LF)", color = "Time Pressure",
       title = "Frequency Ratio 1:4") +
  theme_fig

p_combined_exp1 <- p_12 + p_14
p_combined_exp1

ggsave(file.path(fig_path, "empirical_results.pdf"), p_combined_exp1,
       width = 10, height = 4.5, dpi = 300)
```

**Time pressure x distance**: $\beta$ = `r round(tp_coef_exp1["Estimate"], 3)`, *z* = `r round(tp_coef_exp1["z value"], 2)`, *p* `r ifelse(tp_coef_exp1["Pr(>|z|)"] < .001, "< .001", sprintf("= %.3f", tp_coef_exp1["Pr(>|z|)"]))`

**Frequency ratio x distance**: $\beta$ = `r round(freq_coef_exp1["Estimate"], 3)`, *z* = `r round(freq_coef_exp1["z value"], 2)`, *p* `r ifelse(freq_coef_exp1["Pr(>|z|)"] < .001, "< .001", sprintf("= %.3f", freq_coef_exp1["Pr(>|z|)"]))`

# Response Time Analysis

```{r rt-by-condition}
# Mean RT by condition (2x2)
rt_by_condition <- d_critical_exp1 |>
  group_by(time_pressure, frequency_ratio) |>
  summarise(
    M_rt = mean(rt) / 1000,
    SD_rt = sd(rt) / 1000,
    n = n(),
    .groups = "drop"
  )

# Fitted T values from race model (see models/fit.py)
fitted_T <- tibble(
  time_pressure = c(5, 5, 10, 10),
  frequency_ratio = c("1:2", "1:4", "1:2", "1:4"),
  T_fitted = c(3.47, 5.00, 5.66, 5.66)  # Interaction model: T_5=3.47, T_10=5.66, γ=1.53
)

# Compare empirical RT to fitted T
rt_comparison <- rt_by_condition |>
  left_join(fitted_T, by = c("time_pressure", "frequency_ratio")) |>
  mutate(
    ratio = T_fitted / M_rt
  )

knitr::kable(
  rt_comparison,
  digits = 2,
  col.names = c("Time Pressure (s)", "Freq Ratio", "Mean RT (s)", "SD RT (s)", "N", "Fitted T", "T/RT Ratio"),
  caption = "Comparison of empirical mean RTs and fitted T parameters from race model"
)
```

```{r rt-interaction-test}
# Test for frequency × time pressure interaction in RTs
rt_summary_wide <- rt_by_condition |>
  select(time_pressure, frequency_ratio, M_rt) |>
  pivot_wider(names_from = c(frequency_ratio, time_pressure), values_from = M_rt)

# Compute interaction
delta_rt_12 <- rt_summary_wide$`1:2_10` - rt_summary_wide$`1:2_5`
delta_rt_14 <- rt_summary_wide$`1:4_10` - rt_summary_wide$`1:4_5`

cat("Time pressure effect on RT (ΔRT = RT_10s - RT_5s):\n")
cat(sprintf("  1:2: ΔRT = %.2f s\n", delta_rt_12))
cat(sprintf("  1:4: ΔRT = %.2f s\n", delta_rt_14))
cat(sprintf("  Ratio: %.2fx\n", delta_rt_12 / delta_rt_14))
```

```{r rt-by-response}
# RT by response type (HF vs LF) and condition
d_critical_exp1 |>
  group_by(time_pressure, frequency_ratio, responseFreq) |>
  summarise(
    M = mean(rt) / 1000,
    SD = sd(rt) / 1000,
    n = n(),
    .groups = "drop"
  ) |>
  knitr::kable(
    digits = 2,
    col.names = c("Time Pressure", "Freq Ratio", "Response", "M RT (s)", "SD (s)", "N")
  )
```

```{r rt-model}
m_rt_exp1 <- lmer(
  log(rt) ~ responded_lf * dist_to_lf_scaled * time_pressure_c + freq_ratio_c +
    (1 + dist_to_lf_scaled | subject_id) + (1 | lf_word),
  data = d_critical_exp1,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)
summary(m_rt_exp1)
```

# Summary

```{r summary}
cat("=== EXPERIMENT 1 SUMMARY ===\n\n")
cat("N =", n_participants, "participants\n")
cat("N =", nrow(d_critical_exp1), "critical trials\n")
cat("N =", n_distinct(d_critical_exp1$lf_word), "unique LF words\n\n")
cat("Timeout rates:\n")
cat(sprintf("  Strict (5s): %.1f%%\n", timeout_by_tp_exp1$pct_timeout[1]))
cat(sprintf("  Lenient (10s): %.1f%%\n", timeout_by_tp_exp1$pct_timeout[2]))
cat("\nKey interactions (scaled distance):\n")
cat(sprintf("  Time pressure x distance: B = %.3f, z = %.2f, p %s\n", 
            tp_coef_exp1["Estimate"], tp_coef_exp1["z value"],
            ifelse(tp_coef_exp1["Pr(>|z|)"] < .001, "< .001", sprintf("= %.3f", tp_coef_exp1["Pr(>|z|)"]))))
cat(sprintf("  Frequency ratio x distance: B = %.3f, z = %.2f, p %s\n", 
            freq_coef_exp1["Estimate"], freq_coef_exp1["z value"],
            ifelse(freq_coef_exp1["Pr(>|z|)"] < .001, "< .001", sprintf("= %.3f", freq_coef_exp1["Pr(>|z|)"]))))
```

# Supplementary: Response Proportions by Target Frequency

This analysis examines the proportion of HF, LF, and Other responses as a function of which frequency word is nearest to the target angle. "Other" responses are valid trained words that are neither the nearest nor next-nearest option. Restricted to ambiguous trials (>11° from nearest trained angle) to focus on cases where the target is clearly between two options.

```{r response-proportions-exp1, fig.width=8, fig.height=6}
# Add response category: nearest, next-nearest, or other valid word
d_exp1_with_category <- d_exp1 |>
  filter(between_freqs, ambig_trial, timed_out == 0, !is.na(responseFreq)) |>
  mutate(
    response_category = case_when(
      response_text == nearestLabel ~ nearestFreq,
      response_text == nextNearestLabel ~ nextNearestFreq,
      TRUE ~ "Other"
    )
  )

# Compute response proportions on ambiguous between-frequency trials (aggregate)
d_props_exp1 <- d_exp1_with_category |>
  group_by(nearestFreq, frequency_ratio, time_pressure) |>
  summarise(
    n = n(),
    n_lf = sum(response_category == "LF"),
    n_hf = sum(response_category == "HF"),
    n_other = sum(response_category == "Other"),
    prop_lf = n_lf / n,
    prop_hf = n_hf / n,
    prop_other = n_other / n,
    .groups = "drop"
  )

# By-participant proportions for individual points
d_props_by_subj_exp1 <- d_exp1_with_category |>
  group_by(subject_id, nearestFreq, frequency_ratio, time_pressure) |>
  summarise(
    n = n(),
    prop_lf = sum(response_category == "LF") / n,
    prop_hf = sum(response_category == "HF") / n,
    prop_other = sum(response_category == "Other") / n,
    .groups = "drop"
  ) |>
  pivot_longer(cols = c(prop_lf, prop_hf, prop_other), 
               names_to = "response_type", 
               values_to = "proportion") |>
  mutate(
    response_type = case_when(
      response_type == "prop_lf" ~ "LF",
      response_type == "prop_hf" ~ "HF",
      response_type == "prop_other" ~ "Other"
    ),
    response_type = factor(response_type, levels = c("HF", "LF", "Other")),
    time_pressure = factor(time_pressure, labels = c("5s (Strict)", "10s (Lenient)"))
  )

# Reshape aggregate for plotting
d_props_long_exp1 <- d_props_exp1 |>
  pivot_longer(cols = c(prop_lf, prop_hf, prop_other), 
               names_to = "response_type", 
               values_to = "proportion") |>
  mutate(
    response_type = case_when(
      response_type == "prop_lf" ~ "LF",
      response_type == "prop_hf" ~ "HF",
      response_type == "prop_other" ~ "Other"
    ),
    response_type = factor(response_type, levels = c("HF", "LF", "Other")),
    time_pressure = factor(time_pressure, labels = c("5s (Strict)", "10s (Lenient)"))
  )

# Plot: Response proportions by nearest frequency with participant points
ggplot(d_props_long_exp1, aes(x = nearestFreq, y = proportion, fill = response_type)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.6) +
  geom_point(data = d_props_by_subj_exp1,
             aes(x = nearestFreq, y = proportion, color = response_type),
             position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.9),
             alpha = 0.4, size = 1.5, show.legend = FALSE) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray50") +
  facet_grid(time_pressure ~ frequency_ratio, labeller = labeller(
    frequency_ratio = c("1:2" = "Freq Ratio 1:2", "1:4" = "Freq Ratio 1:4")
  )) +
  scale_fill_manual(values = c("HF" = "#E94F37", "LF" = "#2E86AB", "Other" = "#7D7D7D")) +
  scale_color_manual(values = c("HF" = "#B33D2B", "LF" = "#1F5F7A", "Other" = "#4D4D4D")) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.25)) +
  labs(
    x = "Nearest Trained Angle Frequency",
    y = "Proportion of Responses",
    fill = "Response",
    title = "Exp 1 Response Proportions",
    subtitle = "Bars = group means; Points = individual participants; Other = valid word but not nearest/next-nearest"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold")
  )
```

```{r response-table-exp1}
# Summary table
d_props_exp1 |>
  mutate(
    time_pressure = factor(time_pressure, labels = c("5s", "10s"))
  ) |>
  knitr::kable(
    digits = 2,
    col.names = c("Nearest Freq", "Freq Ratio", "Time Pressure", "N Trials", 
                  "N LF", "N HF", "N Other", "P(LF)", "P(HF)", "P(Other)"),
    caption = "Response proportions on ambiguous between-frequency trials"
  )
```

# Exploratory: Learning Quality Checks

This section examines accuracy on exact-angle trials (timed production at trained angles) and untimed final recall trials to assess overall learning quality.

```{r load-accuracy-data-exp1}
# Need trial_type to identify exact_angle trials, and response to identify final-recall
cols_accuracy <- c("subject_id", "frequency_ratio", "time_pressure", "trial_type",
                   "is_correct", "response", "target", "angle", "targetFreq")

accuracy_data_exp1 <- map_dfr(raw_files, \(f) {
  read_csv(f, show_col_types = FALSE) |>
    select(any_of(cols_accuracy)) |>
    mutate(across(everything(), as.character))
})

# Extract exact_angle trials
exact_angle_exp1 <- accuracy_data_exp1 |>
  filter(trial_type == "exact_angle") |>
  transmute(
    subject_id,
    is_correct = as.numeric(is_correct),
    target,
    angle = as.numeric(angle),
    targetFreq
  ) |>
  left_join(participants, by = "subject_id")

# Extract final recall trials (untimed)
final_recall_exp1 <- accuracy_data_exp1 |>
  filter(str_detect(response, "final-recall-response")) |>
  transmute(
    subject_id,
    is_correct = as.numeric(is_correct),
    target,
    angle = as.numeric(angle),
    targetFreq
  ) |>
  left_join(participants, by = "subject_id")

cat("Exact angle trials:", nrow(exact_angle_exp1), "\n")
cat("Final recall trials:", nrow(final_recall_exp1), "\n")
```



```{r exact-angle-by-targetfreq-exp1, fig.width=8, fig.height=6}
# Exact angle trials: accuracy by target frequency (HF vs LF)
exact_acc_by_freq_exp1 <- exact_angle_exp1 |>
  group_by(targetFreq, frequency_ratio, time_pressure) |>
  summarise(
    n_trials = n(),
    accuracy = mean(is_correct, na.rm = TRUE),
    .groups = "drop"
  )

# By-participant accuracy for individual points
exact_acc_by_freq_subj_exp1 <- exact_angle_exp1 |>
  group_by(subject_id, targetFreq, frequency_ratio, time_pressure) |>
  summarise(
    accuracy = mean(is_correct, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(exact_acc_by_freq_exp1, aes(x = targetFreq, y = accuracy, fill = targetFreq)) +
  geom_col(alpha = 0.6) +
  geom_point(data = exact_acc_by_freq_subj_exp1,
             aes(x = targetFreq, y = accuracy, color = targetFreq),
             position = position_jitter(width = 0.2),
             alpha = 0.4, size = 1.5, show.legend = FALSE) +
  facet_grid(time_pressure ~ frequency_ratio, 
             labeller = labeller(
               time_pressure = c("5" = "5s (Strict)", "10" = "10s (Lenient)"),
               frequency_ratio = c("1:2" = "Freq Ratio 1:2", "1:4" = "Freq Ratio 1:4")
             )) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = c("HF" = "#E94F37", "LF" = "#2E86AB")) +
  scale_color_manual(values = c("HF" = "#B33D2B", "LF" = "#1F5F7A")) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.25)) +
  labs(
    x = "Target Frequency",
    y = "Accuracy",
    fill = "Target Freq",
    title = "Exp 1: Exact Angle Trial Accuracy by Target Frequency",
    subtitle = "Timed production at trained angles; Bars = group means; Points = individual participants"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold")
  )
```

```{r final-recall-by-targetfreq-exp1, fig.width=8, fig.height=6}
# Final recall trials: accuracy by target frequency (HF vs LF)
recall_acc_by_freq_exp1 <- final_recall_exp1 |>
  group_by(targetFreq, frequency_ratio, time_pressure) |>
  summarise(
    n_trials = n(),
    accuracy = mean(is_correct, na.rm = TRUE),
    .groups = "drop"
  )

# By-participant accuracy for individual points
recall_acc_by_freq_subj_exp1 <- final_recall_exp1 |>
  group_by(subject_id, targetFreq, frequency_ratio, time_pressure) |>
  summarise(
    accuracy = mean(is_correct, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(recall_acc_by_freq_exp1, aes(x = targetFreq, y = accuracy, fill = targetFreq)) +
  geom_col(alpha = 0.6) +
  geom_point(data = recall_acc_by_freq_subj_exp1,
             aes(x = targetFreq, y = accuracy, color = targetFreq),
             position = position_jitter(width = 0.2),
             alpha = 0.4, size = 1.5, show.legend = FALSE) +
  facet_grid(time_pressure ~ frequency_ratio, 
             labeller = labeller(
               time_pressure = c("5" = "5s (Strict)", "10" = "10s (Lenient)"),
               frequency_ratio = c("1:2" = "Freq Ratio 1:2", "1:4" = "Freq Ratio 1:4")
             )) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = c("HF" = "#E94F37", "LF" = "#2E86AB")) +
  scale_color_manual(values = c("HF" = "#B33D2B", "LF" = "#1F5F7A")) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.25)) +
  labs(
    x = "Target Frequency",
    y = "Accuracy",
    fill = "Target Freq",
    title = "Exp 1: Final Recall Accuracy by Target Frequency",
    subtitle = "Untimed production at trained angles; Bars = group means; Points = individual participants"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold")
  )
```

```{r accuracy-summary-exp1}
# Compute overall accuracy
exact_acc_overall_exp1 <- mean(exact_angle_exp1$is_correct, na.rm = TRUE)
recall_acc_overall_exp1 <- mean(final_recall_exp1$is_correct, na.rm = TRUE)

cat("\n=== ACCURACY SUMMARY (Exp 1) ===\n")
cat(sprintf("Exact Angle (Timed): %.1f%%\n", 100 * exact_acc_overall_exp1))
cat(sprintf("Final Recall (Untimed): %.1f%%\n", 100 * recall_acc_overall_exp1))
```
