---
title: "Experiment 1: Cost Pressure"
author: "Elves Project Team"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 8
    fig-height: 5
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(lme4)
library(lmerTest)
library(here)
library(patchwork)

# Color palettes
colors_tp <- c("5" = "#E94F37", "10" = "#2E86AB")
colors_freq <- c("1:2" = "#6B4C9A", "1:4" = "#2E86AB")

# Output paths
fig_path <- here("latex/figures")
```

# Data Preprocessing

```{r load-data}
# Load all raw data once, selecting only columns we need to avoid type conflicts
raw_data_path <- here("data/full-production/raw_data/run-1")
raw_files <- list.files(raw_data_path, full.names = TRUE) |>
  str_subset("inprog", negate = TRUE)

cols_needed <- c("subject_id", "frequency_ratio", "time_pressure", "survey_difficulty",
                 "is_correct", "response", "response_text", "target", "angle", "targetFreq",
                 "rt", "nearest_trained_angle", "timed_out")

raw_data <- map_dfr(raw_files, \(f) {
  read_csv(f, show_col_types = FALSE) |>
    select(any_of(cols_needed)) |>
    mutate(across(everything(), as.character))  # coerce all to character, convert later
})

cat("Loaded", n_distinct(raw_data$subject_id), "participants\n")
```

```{r extract-data}
# Extract participant conditions
participants <- raw_data |>
  filter(!is.na(survey_difficulty)) |>
  distinct(subject_id, frequency_ratio, time_pressure) |>
  mutate(time_pressure = as.numeric(time_pressure))

# Extract angle-word mappings from exposure phase
angle_mappings <- raw_data |>
  filter(!is.na(is_correct), str_detect(response, "exposure-response")) |>
  filter(is_correct == "1") |>
  distinct(subject_id, angle = as.numeric(angle), target, targetFreq)

# Extract production trials
production <- raw_data |>
  filter(!is.na(nearest_trained_angle)) |>
  transmute(
    subject_id,
    rt = as.numeric(rt),
    response_text, target,
    angle = as.numeric(angle),
    targetFreq,
    nearest_trained_angle = as.numeric(nearest_trained_angle),
    timed_out = as.numeric(timed_out),
    distance = nearest_trained_angle - angle,
    # Next-nearest is 45° away, opposite direction from nearest
    next_nearest_angle = (nearest_trained_angle + ifelse(distance > 0, -45, 45)) %% 360
  )
```

```{r join-frequencies}
# Build lookup table for angle -> frequency mapping
angle_freq_lookup <- angle_mappings |>
  select(subject_id, angle, freq = targetFreq, label = target)

# Join frequency labels for nearest, next-nearest, and response
d <- production |>
  left_join(angle_freq_lookup |> rename(nearestAngle = angle, nearestFreq = freq),
            by = c("subject_id", "nearest_trained_angle" = "nearestAngle")) |>
  left_join(angle_freq_lookup |> rename(nextNearestFreq = freq, nextNearestLabel = label),
            by = c("subject_id", "next_nearest_angle" = "angle")) |>
  left_join(angle_freq_lookup |> select(subject_id, response_text = label, responseFreq = freq),
            by = c("subject_id", "response_text")) |>
  left_join(participants, by = "subject_id") |>
  mutate(
    between_freqs = nearestFreq != nextNearestFreq,
    # Distance to LF word (key predictor)
    dist_to_lf = ifelse(
      nearestFreq == "LF",
      abs(distance),
      pmin(abs(angle - next_nearest_angle), 360 - abs(angle - next_nearest_angle))
    )
  )

cat("Total trials:", nrow(d), "\n")
cat("Between-frequency trials:", sum(d$between_freqs, na.rm = TRUE), "\n")
```

# Sample Descriptives

```{r sample-descriptives}
n_participants <- n_distinct(participants$subject_id)

participants |>
  count(time_pressure, frequency_ratio) |>
  pivot_wider(names_from = frequency_ratio, values_from = n) |>
  knitr::kable(col.names = c("Time Pressure (s)", "1:2", "1:4"),
               caption = sprintf("Participants per condition (N = %d)", n_participants))
```

```{r timeout-rates}
timeout_by_tp <- d |>
  group_by(time_pressure) |>
  summarise(
    n_trials = n(),
    n_timeout = sum(timed_out, na.rm = TRUE),
    pct_timeout = 100 * n_timeout / n_trials,
    .groups = "drop"
  )

knitr::kable(timeout_by_tp, digits = 1,
             col.names = c("Time Pressure (s)", "N Trials", "N Timeouts", "% Timeout"))
```

# Critical Trial Selection

```{r filter-critical}
d_critical <- d |>
  filter(between_freqs, timed_out == 0, !is.na(responseFreq)) |>
  mutate(
    responded_lf = as.numeric(responseFreq == "LF"),
    time_pressure_c = ifelse(time_pressure == 5, -0.5, 0.5),
    freq_ratio_c = ifelse(frequency_ratio == "1:2", -0.5, 0.5)
  )

cat("Critical trials:", nrow(d_critical), "\n")
cat("N participants:", n_distinct(d_critical$subject_id), "\n")
cat("Overall P(LF):", round(mean(d_critical$responded_lf), 3), "\n")
```

# Figure 2: Empirical Results

```{r figure-helper}
# Helper function for binned plots with model predictions
plot_by_condition <- function(data, condition_var, condition_contrast,
                              colors, legend_title, legend_labels = waiver()) {
  n_bins <- 9

  # Binned means
  d_binned <- data |>
    mutate(dist_bin = cut(dist_to_lf, breaks = n_bins, include.lowest = TRUE)) |>
    group_by(condition = .data[[condition_var]], dist_bin) |>
    summarise(
      n = n(),
      p_lf = mean(responded_lf),
      se = sqrt(p_lf * (1 - p_lf) / n),
      dist_mean = mean(dist_to_lf),
      .groups = "drop"
    )

  # Model predictions
  formula <- as.formula(paste0(
    "responded_lf ~ dist_to_lf * ", condition_contrast, " + (1 | subject_id)"
  ))
  model <- glmer(formula, data = data, family = binomial)

  pred_grid <- expand.grid(
    dist_to_lf = seq(0, 45, length.out = 100),
    contrast = c(-0.5, 0.5)
  )
  names(pred_grid)[2] <- condition_contrast
  pred_grid$p_lf_pred <- predict(model, newdata = pred_grid, type = "response", re.form = NA)
  pred_grid$condition <- levels(factor(data[[condition_var]]))[as.numeric(pred_grid[[condition_contrast]] > 0) + 1]

  ggplot(d_binned, aes(x = dist_mean, y = p_lf, color = factor(condition))) +
    geom_point(size = 3) +
    geom_errorbar(aes(ymin = p_lf - se, ymax = p_lf + se), width = 1) +
    geom_line(data = pred_grid, aes(x = dist_to_lf, y = p_lf_pred), linewidth = 1.2) +
    geom_vline(xintercept = 22.5, linetype = "dashed", color = "gray50") +
    scale_color_manual(values = colors, labels = legend_labels) +
    scale_x_continuous(limits = c(0, 45), breaks = seq(0, 45, by = 10)) +
    scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.25)) +
    labs(x = "Distance from LF (deg)", y = "P(LF)", color = legend_title) +
    theme_minimal(base_size = 12) +
    theme(
      legend.position = c(0.75, 0.85),
      legend.background = element_rect(fill = "white", color = NA),
      panel.grid.minor = element_blank()
    )
}
```

```{r fig-results, fig.width=10, fig.height=4}
#| fig-cap: "P(LF) by distance from LF prototype, split by (A) time pressure and (B) frequency ratio"

p_tp <- plot_by_condition(
  d_critical, "time_pressure", "time_pressure_c",
  colors_tp, "Time Pressure", c("Strict (5s)", "Lenient (10s)")
)

p_freq <- plot_by_condition(
  d_critical, "frequency_ratio", "freq_ratio_c",
  colors_freq, "Frequency Ratio"
)

p_combined <- (p_tp + labs(title = "A")) + (p_freq + labs(title = "B")) &
  theme(plot.title = element_text(face = "bold"))

p_combined

ggsave(file.path(fig_path, "empirical_results.pdf"), p_combined,
       width = 10, height = 4, dpi = 300)
```

# Statistical Models

```{r model-tp}
m_tp <- glmer(
  responded_lf ~ time_pressure_c * dist_to_lf + freq_ratio_c + (1 | subject_id),
  data = d_critical, family = binomial
)
summary(m_tp)
```

```{r model-freq}
m_freq <- glmer(
  responded_lf ~ freq_ratio_c * dist_to_lf + time_pressure_c + (1 | subject_id),
  data = d_critical, family = binomial
)
summary(m_freq)
```

```{r extract-coefficients}
tp_coef <- summary(m_tp)$coefficients["time_pressure_c:dist_to_lf", ]
freq_coef <- summary(m_freq)$coefficients["freq_ratio_c:dist_to_lf", ]
```

**Time pressure x distance**: $\beta$ = `r round(tp_coef["Estimate"], 3)`, *z* = `r round(tp_coef["z value"], 2)`, *p* < .001

**Frequency ratio x distance**: $\beta$ = `r round(freq_coef["Estimate"], 3)`, *z* = `r round(freq_coef["z value"], 2)`, *p* < .001

# Response Time Analysis

```{r rt-summary}
d_critical |>
  group_by(time_pressure, responseFreq) |>
  summarise(M = mean(rt), SD = sd(rt), n = n(), .groups = "drop") |>
  knitr::kable(digits = 0, col.names = c("Time Pressure", "Response", "M RT", "SD", "N"))
```

```{r rt-model}
m_rt <- lmer(
  log(rt) ~ responded_lf * dist_to_lf * time_pressure_c + (1 | subject_id),
  data = d_critical
)
summary(m_rt)
```

# Summary

```{r summary}
cat("=== EXPERIMENT 1 SUMMARY ===\n\n")
cat("N =", n_participants, "participants\n")
cat("N =", nrow(d_critical), "critical trials\n\n")
cat("Timeout rates:\n")
cat(sprintf("  Strict (5s): %.1f%%\n", timeout_by_tp$pct_timeout[1]))
cat(sprintf("  Lenient (10s): %.1f%%\n", timeout_by_tp$pct_timeout[2]))
cat("\nKey interactions:\n")
cat(sprintf("  Time pressure x distance: B = %.3f, z = %.2f\n", tp_coef["Estimate"], tp_coef["z value"]))
cat(sprintf("  Frequency ratio x distance: B = %.3f, z = %.2f\n", freq_coef["Estimate"], freq_coef["z value"]))
```
