---
title: "Experiment 1: Cost Pressure"
author: "Elves Project Team"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 8
    fig-height: 5
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(lme4)
library(lmerTest)
library(here)
library(patchwork)

# Color palette
colors_tp <- c("5" = "#E94F37", "10" = "#2E86AB")
colors_freq <- c("1:2" = "#6B4C9A", "1:4" = "#2E86AB")

# Output paths
fig_path <- here("latex/figures")
```

# Data Preprocessing

Load and clean raw experimental data.

```{r load-raw-data}
raw_data_path <- here("data/full-production/raw_data/run-1")
raw_files <- grep("inprog", list.files(raw_data_path, full.names = TRUE),
                  value = TRUE, invert = TRUE)

cat("Found", length(raw_files), "participant data files\n")
```

## Extract participant-level data

```{r extract-participants}
extract_participant_data <- function(raw_data_file) {
  read_csv(raw_data_file, show_col_types = FALSE) |>
    select(subject_id, prolific_id, frequency_ratio, time_pressure,
           selected_training_block_id, selected_test_block_id,
           survey_difficulty) |>  # Only select columns we need
    filter(!is.na(survey_difficulty))
}

participants <- map_dfr(raw_files, extract_participant_data) |>
  select(-prolific_id)  # Remove PII

cat("N participants:", nrow(participants), "\n")
```

## Extract exposure data (for angle-word mappings)

```{r extract-exposure}
extract_exposure_data <- function(raw_data_file) {
  read_csv(raw_data_file, show_col_types = FALSE) |>
    select(subject_id, is_correct, response_text, target, angle, targetFreq, response) |>
    filter(!is.na(is_correct)) |>
    mutate(parsed = map(response, jsonlite::parse_json)) |>
    unnest_wider(parsed, names_sep = "_") |>
    filter(!is.na(`parsed_exposure-response`)) |>
    select(subject_id, is_correct, response_text, target, angle, targetFreq)
}

exposure <- map_dfr(raw_files, extract_exposure_data)

# Store angle-word mappings per participant
angle_mappings <- exposure |>
  filter(is_correct == 1) |>
  select(subject_id, target, angle, targetFreq)
```

## Extract and clean production data

```{r extract-production}
extract_production_data <- function(raw_data_file) {
  read_csv(raw_data_file, show_col_types = FALSE) |>
    select(subject_id, rt, is_correct, response_text, target, angle, targetFreq,
           trial_id, nearest_trained_angle, bonus_points, total_points, timed_out) |>
    filter(!is.na(nearest_trained_angle))
}

production_raw <- map_dfr(raw_files, extract_production_data)

# Clean production data: fix angle assignments and join frequency labels
production <- production_raw |>
  mutate(
    distance = nearest_trained_angle - angle,
    # Compute next-nearest angle (45° away from nearest)
    c_next_nearest = ifelse(distance > 0,
                            nearest_trained_angle - 45,
                            nearest_trained_angle + 45),
    next_nearest_angle = case_when(
      c_next_nearest < 0 ~ 360 + c_next_nearest,
      c_next_nearest > 360 ~ c_next_nearest - 360,
      TRUE ~ c_next_nearest
    )
  ) |>
  select(
    subject_id, rt, is_correct, trial_id, response_text,
    targetAngle = angle, targetLabel = target, targetFreq,
    nearestAngle = nearest_trained_angle,
    nextNearestAngle = next_nearest_angle,
    distance, bonus_points, total_points, timed_out
  ) |>
  # Join frequency labels from angle mappings
  left_join(
    angle_mappings |>
      rename(nextNearestAngle = angle, nextNearestFreq = targetFreq,
             nextNearestLabel = target),
    by = c("subject_id", "nextNearestAngle")
  ) |>
  left_join(
    angle_mappings |>
      select(-target) |>
      rename(nearestAngle = angle, nearestFreq = targetFreq),
    by = c("subject_id", "nearestAngle")
  ) |>
  left_join(
    angle_mappings |>
      select(subject_id, response_text = target, responseFreq = targetFreq),
    by = c("subject_id", "response_text")
  ) |>
  # Compute trial type flags

mutate(
    between_freqs = (nearestFreq == "HF" & nextNearestFreq == "LF") |
                    (nearestFreq == "LF" & nextNearestFreq == "HF"),
    ambig_trial = distance <= -11,
    is_critical = between_freqs & ambig_trial
  ) |>
  # Compute distance to LF word (key predictor)
  mutate(
    raw_dist_to_lf = ifelse(nearestFreq == "LF",
                            abs(distance),
                            abs(targetAngle - nextNearestAngle)),
    # Handle circular wraparound
    dist_to_lf = ifelse(raw_dist_to_lf > 180, 360 - raw_dist_to_lf, raw_dist_to_lf)
  ) |>
  select(-raw_dist_to_lf)

cat("Production trials:", nrow(production), "\n")
cat("Between-frequency trials:", sum(production$between_freqs, na.rm = TRUE), "\n")
```

## Merge with participant conditions

```{r merge-data}
d <- production |>
  left_join(
    participants |> select(subject_id, frequency_ratio, time_pressure),
    by = "subject_id"
  )
```

# Sample Descriptives

```{r sample-descriptives}
n_participants <- n_distinct(participants$subject_id)

condition_counts <- participants |>
  count(time_pressure, frequency_ratio) |>
  pivot_wider(names_from = frequency_ratio, values_from = n)

cat("Total participants:", n_participants, "\n\n")
knitr::kable(condition_counts, col.names = c("Time Pressure (s)", "1:2", "1:4"),
             caption = "Participants per condition")
```

# Timeout Rates

```{r timeout-rates}
timeout_by_tp <- d |>
  group_by(time_pressure) |>
  summarise(
    n_trials = n(),
    n_timeout = sum(timed_out, na.rm = TRUE),
    pct_timeout = 100 * n_timeout / n_trials,
    .groups = "drop"
  )

knitr::kable(timeout_by_tp, digits = 1,
             col.names = c("Time Pressure (s)", "N Trials", "N Timeouts", "% Timeout"))
```

# Critical Trial Selection

Critical trials: target falls between LF and HF trained directions.

```{r filter-critical}
d_critical <- d |>
  filter(
    between_freqs == TRUE,
    timed_out == 0,
    !is.na(responseFreq)
  ) |>
  mutate(
    responded_lf = as.numeric(responseFreq == "LF"),
    time_pressure_c = ifelse(time_pressure == 5, -0.5, 0.5),
    freq_ratio_c = ifelse(frequency_ratio == "1:2", -0.5, 0.5)
  )

cat("Critical trials:\n")
cat("  N participants:", n_distinct(d_critical$subject_id), "\n")
cat("  N trials:", nrow(d_critical), "\n")
cat("  Overall P(LF):", round(mean(d_critical$responded_lf), 3), "\n")
cat("  dist_to_lf range:", range(d_critical$dist_to_lf), "degrees\n")
```

# Descriptive Statistics

```{r descriptives}
desc_stats <- d_critical |>
  group_by(time_pressure, frequency_ratio) |>
  summarise(
    n = n(),
    p_lf = mean(responded_lf),
    sd = sd(responded_lf),
    .groups = "drop"
  )

knitr::kable(desc_stats, digits = 3,
             col.names = c("Time Pressure", "Freq Ratio", "N", "P(LF)", "SD"))
```

# Figure 2: Empirical Results

## Panel A: Time Pressure Effect

```{r fig-time-pressure, fig.width=5, fig.height=4}
# Binned summary
n_bins <- 9
d_binned_tp <- d_critical |>
  mutate(dist_bin = cut(dist_to_lf, breaks = n_bins, include.lowest = TRUE)) |>
  group_by(time_pressure, dist_bin) |>
  summarise(
    n = n(),
    p_lf = mean(responded_lf),
    se = sqrt(p_lf * (1 - p_lf) / n),
    dist_mean = mean(dist_to_lf),
    .groups = "drop"
  )

# Model for smooth curves
m_tp_plot <- glmer(
  responded_lf ~ dist_to_lf * time_pressure_c + (1 | subject_id),
  data = d_critical,
  family = binomial(link = "logit")
)

# Predictions
pred_tp <- expand.grid(
  dist_to_lf = seq(0, 45, length.out = 100),
  time_pressure_c = c(-0.5, 0.5)
) |>
  mutate(
    time_pressure = ifelse(time_pressure_c == -0.5, 5, 10),
    p_lf_pred = predict(m_tp_plot, newdata = pick(everything()),
                        type = "response", re.form = NA)
  )

# Plot
p_tp <- ggplot(d_binned_tp, aes(x = dist_mean, y = p_lf,
                                 color = factor(time_pressure))) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = p_lf - se, ymax = p_lf + se), width = 1) +
  geom_line(data = pred_tp, aes(x = dist_to_lf, y = p_lf_pred), linewidth = 1.2) +
  geom_vline(xintercept = 22.5, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = colors_tp, labels = c("Strict (5s)", "Lenient (10s)")) +
  scale_x_continuous(limits = c(0, 45), breaks = seq(0, 45, by = 10)) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.25)) +
  labs(x = "Distance from LF (°)", y = "P(LF)", color = "Time Pressure") +
  theme_minimal(base_size = 12) +
  theme(legend.position = c(0.75, 0.85),
        legend.background = element_rect(fill = "white", color = NA),
        panel.grid.minor = element_blank())

p_tp
```

## Panel B: Frequency Ratio Effect

```{r fig-freq-ratio, fig.width=5, fig.height=4}
# Binned summary
d_binned_freq <- d_critical |>
  mutate(dist_bin = cut(dist_to_lf, breaks = n_bins, include.lowest = TRUE)) |>
  group_by(frequency_ratio, dist_bin) |>
  summarise(
    n = n(),
    p_lf = mean(responded_lf),
    se = sqrt(p_lf * (1 - p_lf) / n),
    dist_mean = mean(dist_to_lf),
    .groups = "drop"
  )

# Model for smooth curves
m_freq_plot <- glmer(
  responded_lf ~ dist_to_lf * freq_ratio_c + (1 | subject_id),
  data = d_critical,
  family = binomial(link = "logit")
)

# Predictions
pred_freq <- expand.grid(
  dist_to_lf = seq(0, 45, length.out = 100),
  freq_ratio_c = c(-0.5, 0.5)
) |>
  mutate(
    frequency_ratio = ifelse(freq_ratio_c == -0.5, "1:2", "1:4"),
    p_lf_pred = predict(m_freq_plot, newdata = pick(everything()),
                        type = "response", re.form = NA)
  )

# Plot
p_freq <- ggplot(d_binned_freq, aes(x = dist_mean, y = p_lf, color = frequency_ratio)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = p_lf - se, ymax = p_lf + se), width = 1) +
  geom_line(data = pred_freq, aes(x = dist_to_lf, y = p_lf_pred), linewidth = 1.2) +
  geom_vline(xintercept = 22.5, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = colors_freq) +
  scale_x_continuous(limits = c(0, 45), breaks = seq(0, 45, by = 10)) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.25)) +
  labs(x = "Distance from LF (°)", y = "P(LF)", color = "Frequency Ratio") +
  theme_minimal(base_size = 12) +
  theme(legend.position = c(0.75, 0.85),
        legend.background = element_rect(fill = "white", color = NA),
        panel.grid.minor = element_blank())

p_freq
```

## Combined Figure

```{r fig-combined, fig.width=10, fig.height=4}
p_combined <- (p_tp + labs(title = "A") + theme(plot.title = element_text(face = "bold"))) +
              (p_freq + labs(title = "B") + theme(plot.title = element_text(face = "bold")))

p_combined

# Save for paper
ggsave(file.path(fig_path, "empirical_results.pdf"), p_combined,
       width = 10, height = 4, dpi = 300)
```

# Statistical Models

## Model 1: Main Effects

```{r model-main}
m1 <- glmer(
  responded_lf ~ time_pressure_c + freq_ratio_c + dist_to_lf + (1 | subject_id),
  data = d_critical,
  family = binomial(link = "logit")
)

summary(m1)
```

## Model 2: Time Pressure × Distance Interaction

```{r model-tp-interaction}
m2 <- glmer(
  responded_lf ~ time_pressure_c * dist_to_lf + freq_ratio_c + (1 | subject_id),
  data = d_critical,
  family = binomial(link = "logit")
)

summary(m2)
anova(m1, m2)
```

```{r}
tp_coef <- summary(m2)$coefficients["time_pressure_c:dist_to_lf", ]
```

**Time pressure × distance interaction**: β = `r round(tp_coef["Estimate"], 3)`, z = `r round(tp_coef["z value"], 2)`, p < .001

## Model 3: Frequency Ratio × Distance Interaction

```{r model-freq-interaction}
m3 <- glmer(
  responded_lf ~ freq_ratio_c * dist_to_lf + time_pressure_c + (1 | subject_id),
  data = d_critical,
  family = binomial(link = "logit")
)

summary(m3)
```

```{r}
freq_coef <- summary(m3)$coefficients["freq_ratio_c:dist_to_lf", ]
```

**Frequency ratio × distance interaction**: β = `r round(freq_coef["Estimate"], 3)`, z = `r round(freq_coef["z value"], 2)`, p < .001

## Model 4: Full Three-Way Interaction

```{r model-full}
m4 <- glmer(
  responded_lf ~ time_pressure_c * freq_ratio_c * dist_to_lf + (1 | subject_id),
  data = d_critical,
  family = binomial(link = "logit")
)

summary(m4)
```

# Response Time Analysis

```{r rt-descriptives}
rt_summary <- d_critical |>
  group_by(time_pressure, frequency_ratio) |>
  summarise(
    M_rt = mean(rt),
    SD_rt = sd(rt),
    Median_rt = median(rt),
    .groups = "drop"
  )

knitr::kable(rt_summary, digits = 0,
             col.names = c("Time Pressure", "Freq Ratio", "M RT", "SD", "Median"))
```

```{r rt-by-response}
rt_by_resp <- d_critical |>
  group_by(responseFreq, time_pressure) |>
  summarise(
    M_rt = mean(rt),
    SD_rt = sd(rt),
    n = n(),
    .groups = "drop"
  )

knitr::kable(rt_by_resp, digits = 0,
             col.names = c("Response", "Time Pressure", "M RT", "SD", "N"))
```

```{r rt-model}
d_critical <- d_critical |> mutate(log_rt = log(as.numeric(rt)))

m_rt <- lmer(
  log_rt ~ responded_lf * dist_to_lf * time_pressure_c + (1 | subject_id),
  data = d_critical
)

summary(m_rt)
```

# Summary for Manuscript

```{r summary-for-paper}
cat("=== EXPERIMENT 1 SUMMARY ===\n\n")

cat("Participants: N =", n_participants, "\n")

cat("\nTimeout rates:\n")
cat(sprintf("  Strict (5s): %.1f%%\n",
            timeout_by_tp$pct_timeout[timeout_by_tp$time_pressure == 5]))
cat(sprintf("  Lenient (10s): %.1f%%\n",
            timeout_by_tp$pct_timeout[timeout_by_tp$time_pressure == 10]))

cat("\nTime pressure × distance interaction:\n")
cat(sprintf("  β = %.3f, z = %.2f, p < .001\n",
            tp_coef["Estimate"], tp_coef["z value"]))

cat("\nFrequency ratio × distance interaction:\n
")
cat(sprintf("  β = %.3f, z = %.2f, p < .001\n",
            freq_coef["Estimate"], freq_coef["z value"]))
```

# Session Info

```{r session-info}
sessionInfo()
```
