---
title: "Experiment 2: Benefit Pressure"
author: "Elves Project Team"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 8
    fig-height: 5
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(tidyboot)
library(lme4)
library(lmerTest)
library(here)
library(patchwork)

# Color palette
# Color palettes
colors_tp <- c("5" = "#E94F37", "10" = "#2E86AB")
colors_payoff <- c("linear" = "#6B4C9A", "binary" = "#2E86AB")

# Output paths
fig_path <- here("latex/figures")
```

# Data Preprocessing

```{r load-data}
# Load data from three sources (match models/fit_exp2.py):
# - exp2-0: binary payoff, 1:4 (exclude 1:4 linear from exp2-0 — replaced by exp2-1)
# - exp2-1: linear payoff, 1:4
# - exp2-2: frequency ratio 1:2 (both payoff structures)

cols_needed <- c("subject_id", "frequency_ratio", "time_pressure", "payoff_structure", "survey_difficulty",
                 "is_correct", "response", "response_text", "target", "angle", "targetFreq",
                 "rt", "nearest_trained_angle", "timed_out", "total_points", "prolific_id",
                 "familiarization_block_iteration", "recall_iteration")

# Normalize column names (some CSVs use "Time Pressure" etc.)
normalize_cols <- function(df) {
  nms <- names(df)
  nms_lower <- str_replace_all(str_to_lower(str_trim(nms)), "\\s+", "_")
  canonical <- c("time_pressure", "payoff_structure", "frequency_ratio")
  idx <- match(canonical, nms_lower)
  for (i in seq_along(idx)) {
    if (!is.na(idx[i]) && nms[idx[i]] != canonical[i])
      names(df)[idx[i]] <- canonical[i]
  }
  df
}

load_exp2_folder <- function(folder, datasource_name) {
  path <- here("data", "full-production", "raw_data", folder)
  if (!dir.exists(path)) return(NULL)
  files <- list.files(path, full.names = TRUE) |>
    str_subset("training_complete", negate = TRUE) |>
    str_subset("inprog", negate = TRUE)
  if (length(files) == 0) return(NULL)
  out <- map_dfr(files, \(f) {
    read_csv(f, show_col_types = FALSE) |>
      normalize_cols() |>
      select(any_of(cols_needed)) |>
      mutate(across(everything(), as.character))
  })
  out$datasource <- datasource_name
  out
}

raw_exp2_0 <- load_exp2_folder("exp2-0", "exp2-0")
raw_exp2_1 <- load_exp2_folder("exp2-1", "exp2-1")
raw_exp2_2 <- load_exp2_folder("exp2-2", "exp2-2")

if (!is.null(raw_exp2_0)) cat("Loaded", n_distinct(raw_exp2_0$subject_id), "participants from exp2-0\n")
if (!is.null(raw_exp2_1)) cat("Loaded", n_distinct(raw_exp2_1$subject_id), "participants from exp2-1\n")
if (!is.null(raw_exp2_2)) cat("Loaded", n_distinct(raw_exp2_2$subject_id), "participants from exp2-2\n")

raw_list <- list(raw_exp2_0, raw_exp2_1, raw_exp2_2)
raw_list <- raw_list[!sapply(raw_list, is.null)]
raw_data <- bind_rows(raw_list)
if (nrow(raw_data) == 0) stop("No Exp 2 raw data found under data/full-production/raw_data/exp2-0, exp2-1, exp2-2.")

# Standardize payoff_structure and frequency_ratio
raw_data <- raw_data |>
  mutate(
    payoff_structure = str_trim(str_to_lower(replace_na(payoff_structure, "linear"))),
    payoff_structure = if_else(payoff_structure %in% c("linear", "binary"), payoff_structure, "linear"),
    frequency_ratio = str_trim(replace_na(frequency_ratio, "1:4")),
    frequency_ratio = if_else(frequency_ratio %in% c("1:2", "1:4"), frequency_ratio, "1:4")
  )

# Exclude exp2-0, 1:4, linear (replaced by exp2-1; match fit_exp2.py)
raw_data <- raw_data |>
  filter(!(datasource == "exp2-0" & frequency_ratio == "1:4" & payoff_structure == "linear"))
if ("datasource" %in% names(raw_data)) raw_data <- raw_data |> select(-datasource)

cat("Total:", n_distinct(raw_data$subject_id), "participants\n")

participant_bonus <- raw_data |>
  filter(is.na(prolific_id))
```

```{r avg-training-blocks}
# Average number of training blocks before test phase: for each participant, max block; then mean across participants
avg_training_blocks <- raw_data |>
  filter(!is.na(familiarization_block_iteration)) |>
  mutate(block = as.numeric(familiarization_block_iteration)) |>
  group_by(subject_id) |>
  summarise(max_block = max(block), .groups = "drop") |>
  summarise(mean_blocks = mean(max_block), n = n()) |>
  as.list()
cat("Average training blocks before test: ", round(avg_training_blocks$mean_blocks, 2),
    " (n = ", avg_training_blocks$n, " participants)\n", sep = "")
```

```{r avg-recall-blocks}
# Average number of recall blocks: for each participant, max recall iteration; then mean across participants
avg_recall_blocks <- raw_data |>
  filter(!is.na(recall_iteration)) |>
  mutate(block = as.numeric(recall_iteration)) |>
  group_by(subject_id) |>
  summarise(max_block = max(block), .groups = "drop") |>
  summarise(mean_blocks = mean(max_block), n = n()) |>
  as.list()
cat("Average recall blocks: ", round(avg_recall_blocks$mean_blocks, 2),
    " (n = ", avg_recall_blocks$n, " participants)\n", sep = "")
```

```{r extract-data}
# Extract participant conditions (one row per subject; include frequency_ratio from exp2-0/1/2)
participants <- raw_data |>
  filter(!is.na(time_pressure)) |>
  group_by(subject_id) |>
  summarise(
    time_pressure = as.numeric(first(time_pressure)),
    payoff_structure = first(payoff_structure),
    frequency_ratio = first(frequency_ratio),
    .groups = "drop"
  )

# Extract angle-word mappings from exposure phase (match Python: is_correct & response contains "exposure"; correct = 1 or "1"/"true"/"yes")
angle_mappings <- raw_data |>
  filter(!is.na(is_correct), str_detect(str_to_lower(response), "exposure")) |>
  filter(
    is_correct == "1" |
    replace_na(as.numeric(is_correct), 0) == 1 |
    str_trim(str_to_lower(as.character(is_correct))) %in% c("1", "true", "yes")
  ) |>
  distinct(subject_id, angle = as.numeric(angle), target, targetFreq)

# Extract production trials (match Python: next-nearest reconstructed from geometry, not jsPsych column)
production <- raw_data |>
  filter(!is.na(nearest_trained_angle)) |>
  transmute(
    subject_id,
    rt = as.numeric(rt),
    response_text, target,
    angle = as.numeric(angle),
    targetFreq,
    nearest_trained_angle = as.numeric(nearest_trained_angle),
    timed_out = as.numeric(timed_out)
  ) |>
  mutate(
    # Circular distance to nearest (handles 0°/360° wraparound)
    distance = pmin(abs(angle - nearest_trained_angle), 360 - abs(angle - nearest_trained_angle)),
    # Next-nearest = closer of ±45° neighbors (reconstruct from geometry; do not use jsPsych next_nearest_angle)
    neighbor_plus = (nearest_trained_angle + 45) %% 360,
    neighbor_minus = (nearest_trained_angle - 45 + 360) %% 360,
    dist_plus = pmin(abs(angle - neighbor_plus), 360 - abs(angle - neighbor_plus)),
    dist_minus = pmin(abs(angle - neighbor_minus), 360 - abs(angle - neighbor_minus)),
    next_nearest_angle = ifelse(dist_plus < dist_minus, neighbor_plus, neighbor_minus)
  ) |>
  select(-neighbor_plus, -neighbor_minus, -dist_plus, -dist_minus)
```

```{r join-frequencies}
# Build lookup table for angle -> frequency mapping
angle_freq_lookup <- angle_mappings |>
  select(subject_id, angle, freq = targetFreq, label = target)

# Join frequency labels for nearest, next-nearest, and response
d_exp2 <- production |>
  left_join(angle_freq_lookup |> rename(nearestAngle = angle, nearestFreq = freq, nearestLabel = label),
            by = c("subject_id", "nearest_trained_angle" = "nearestAngle")) |>
  left_join(angle_freq_lookup |> rename(nextNearestFreq = freq, nextNearestLabel = label),
            by = c("subject_id", "next_nearest_angle" = "angle")) |>
  left_join(angle_freq_lookup |> select(subject_id, response_text = label, responseFreq = freq),
            by = c("subject_id", "response_text")) |>
  left_join(participants, by = "subject_id") |>
  mutate(
    between_freqs = nearestFreq != nextNearestFreq,
    # Ambiguous trial: test angle is past midpoint (>11° from nearest, toward next-nearest)
    ambig_trial = distance > 11,
    # Critical trial: between HF/LF frequencies AND in ambiguous zone
    is_critical = between_freqs & ambig_trial,
    # Distance to LF word (key predictor)
    # distance is already circular, and next_nearest uses circular distance
    dist_to_lf = ifelse(
      nearestFreq == "LF",
      distance,
      pmin(abs(angle - next_nearest_angle), 360 - abs(angle - next_nearest_angle))
    ),
    # LF word label for this trial (for random effects)
    lf_word = ifelse(nearestFreq == "LF", nearestLabel, nextNearestLabel)
  )

cat("Total trials:", nrow(d_exp2), "\n")
cat("Between-frequency trials:", sum(d_exp2$between_freqs, na.rm = TRUE), "\n")
cat("Ambiguous trials:", sum(d_exp2$ambig_trial, na.rm = TRUE), "\n")
cat("Critical trials:", sum(d_exp2$is_critical, na.rm = TRUE), "\n")
```

# Sample Descriptives

```{r sample-descriptives}
n_participants <- n_distinct(participants$subject_id)

participants |>
  count(frequency_ratio, time_pressure, payoff_structure)
```

```{r timeout-rates}
# TODO: Timeout rates by condition
```

# Critical Trial Selection

**Match Python (fit_exp2.py):** between_freqs only; timed_out == 0; **200 < rt < time_pressure×1000** (rt in ms, time_pressure in s); no ambig_trial filter. No other editing of trial set.

```{r filter-critical}
timeout_by_tp_exp2 <- d_exp2 |>
  group_by(time_pressure) |>
  summarise(
    n_trials = n(),
    n_timeout = sum(timed_out, na.rm = TRUE),
    pct_timeout = 100 * n_timeout / n_trials,
    .groups = "drop"
  )

knitr::kable(timeout_by_tp_exp2, digits = 1,
             col.names = c("Time Pressure (s)", "N Trials", "N Timeouts", "% Timeout"))
```

# Statistical Models

```{r models}
# Critical trials: same definition as Python (between_freqs; timed_out==0; 200 < rt < time_pressure*1000)
d_critical_exp2 <- d_exp2 |>
  filter(
    between_freqs,
    timed_out == 0,
    !is.na(responseFreq),
    rt > 200,
    rt < time_pressure * 1000   # rt in ms, time_pressure in seconds
  ) |>
  mutate(
    responded_lf = as.numeric(responseFreq == "LF"),
    time_pressure_c = ifelse(time_pressure == 5, -0.5, 0.5),
    payoff_c = ifelse(payoff_structure == "linear", -0.5, 0.5),
    dist_to_lf_scaled = scale(dist_to_lf)[,1]
  )

cat("Critical trials:", nrow(d_critical_exp2), "\n")
cat("N participants:", n_distinct(d_critical_exp2$subject_id), "\n")
cat("N unique LF words:", n_distinct(d_critical_exp2$lf_word), "\n")
cat("Overall P(LF):", round(mean(d_critical_exp2$responded_lf), 3), "\n")
```

```{r model-summaries}
# Time pressure × distance interaction
# Random effects: by-participant intercepts and slopes for distance (uncorrelated), by-word intercepts
m_tp_exp2 <- glmer(
  responded_lf ~ time_pressure_c * dist_to_lf_scaled + payoff_c + 
    (1 + dist_to_lf_scaled | subject_id) + (1 | lf_word),
  data = d_critical_exp2, family = binomial,
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

# Payoff structure × distance interaction
m_payoff_exp2 <- glmer(
  responded_lf ~ payoff_c * dist_to_lf_scaled + time_pressure_c + 
    (1 + dist_to_lf_scaled | subject_id) + (1 | lf_word),
  data = d_critical_exp2, family = binomial,
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

tp_coef_exp2 <- summary(m_tp_exp2)$coefficients["time_pressure_c:dist_to_lf_scaled", ]
payoff_coef_exp2 <- summary(m_payoff_exp2)$coefficients["payoff_c:dist_to_lf_scaled", ]

summary(m_tp_exp2)
summary(m_payoff_exp2)
```

# Model: Chose Nearest by Target Frequency

This analysis examines whether participants chose the nearest trained angle as a function of whether that angle was HF or LF.

```{r model-chose-nearest}
# Prepare data for chose-nearest analysis
d_chose_nearest_exp2 <- d_exp2 |>
  filter(between_freqs, timed_out == 0, !is.na(responseFreq)) |>
  mutate(
    # Outcome: did participant respond with the nearest trained angle's word?
    chose_nearest = as.numeric(response_text == nearestLabel),
    # Predictor: is the nearest angle HF? (centered: -0.5 = LF, 0.5 = HF)
    is_hf_trial_c = ifelse(nearestFreq == "HF", 0.5, -0.5),
    # Centered predictors
    time_pressure_c = ifelse(time_pressure == 5, -0.5, 0.5),
    payoff_c = ifelse(payoff_structure == "linear", -0.5, 0.5),
    distance_c = scale(distance)[,1]
  )

cat("Chose-nearest analysis:\n")
cat("  N trials:", nrow(d_chose_nearest_exp2), "\n")
cat("  N participants:", n_distinct(d_chose_nearest_exp2$subject_id), "\n")
cat("  Overall P(chose nearest):", round(mean(d_chose_nearest_exp2$chose_nearest), 3), "\n")
cat("  P(chose nearest | HF nearest):", round(mean(d_chose_nearest_exp2$chose_nearest[d_chose_nearest_exp2$nearestFreq == "HF"]), 3), "\n")
cat("  P(chose nearest | LF nearest):", round(mean(d_chose_nearest_exp2$chose_nearest[d_chose_nearest_exp2$nearestFreq == "LF"]), 3), "\n")
```

```{r model-chose-nearest-fit}
# Full model: is_hf_trial × payoff × time_pressure interaction + distance
# Design: frequency_ratio (1:2, 1:4) × time_pressure (5s, 10s) × payoff_structure (linear, binary)
m_chose_nearest_exp2 <- glmer(
  chose_nearest ~ is_hf_trial_c * payoff_c * time_pressure_c + distance_c +
    (1 + distance_c | subject_id) + (1 | nearestLabel),
  data = d_chose_nearest_exp2, family = binomial,
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

summary(m_chose_nearest_exp2)
```


```{r chose-nearest-by-distance, fig.width=5, fig.height=4}
# Plot P(chose nearest) by distance, split by HF/LF nearest
n_bins <- 9  # Define here for use in binning

# Filter to 0-22.5 range and bin
d_binned_chose_nearest <- d_chose_nearest_exp2 |>
  filter(distance <= 22.5) |>
  mutate(dist_bin = cut(distance, breaks = n_bins, include.lowest = TRUE)) |>
  group_by(nearestFreq, time_pressure, dist_bin) |>
  tidyboot_mean(chose_nearest) |>
  left_join(
    d_chose_nearest_exp2 |>
      filter(distance <= 22.5) |>
      mutate(dist_bin = cut(distance, breaks = n_bins, include.lowest = TRUE)) |>
      group_by(nearestFreq, time_pressure, dist_bin) |>
      summarise(dist_mean = mean(distance), .groups = "drop"),
    by = c("nearestFreq", "time_pressure", "dist_bin")
  )

# Model predictions for the curve
# Get scaling parameters from the data used to fit the model
dist_mean_cn <- mean(d_chose_nearest_exp2$distance)
dist_sd_cn <- sd(d_chose_nearest_exp2$distance)

pred_chose_nearest <- expand.grid(
  distance = seq(0, 22.5, length.out = 100),
  is_hf_trial_c = c(-0.5, 0.5),
  time_pressure_c = c(-0.5, 0.5),
  payoff_c = 0  # Average across payoff conditions
) |>
  mutate(
    distance_c = (distance - dist_mean_cn) / dist_sd_cn,
    nearestFreq = ifelse(is_hf_trial_c > 0, "HF", "LF"),
    time_pressure = ifelse(time_pressure_c < 0, 5, 10),
    p_chose_nearest = predict(m_chose_nearest_exp2, newdata = pick(everything()), 
                               type = "response", re.form = NA)
  )

ggplot(d_binned_chose_nearest, aes(x = dist_mean, y = empirical_stat, 
                                    color = nearestFreq, shape = factor(time_pressure))) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.5) +
  geom_line(data = pred_chose_nearest, 
            aes(x = distance, y = p_chose_nearest, linetype = factor(time_pressure)),
            linewidth = 1) +
  scale_color_manual(values = c("HF" = "#E94F37", "LF" = "#2E86AB")) +
  scale_linetype_manual(values = c("5" = "solid", "10" = "dashed"),
                        labels = c("5s (Strict)", "10s (Lenient)")) +
  scale_shape_manual(values = c("5" = 16, "10" = 17),
                     labels = c("5s (Strict)", "10s (Lenient)")) +
  scale_x_continuous(limits = c(0, 22.5), breaks = seq(0, 22.5, by = 5)) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.25)) +
  labs(
    x = "Distance from Nearest Trained Angle (deg)",
    y = "P(Chose Nearest)",
    color = "Nearest Freq",
    shape = "Time Pressure",
    linetype = "Time Pressure",
    title = "Exp 2: P(Chose Nearest) by Distance and Target Frequency (Time Pressure)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "right",
    legend.background = element_rect(fill = "white", color = "gray80")
  )
```

```{r chose-nearest-by-distance-payoff, fig.width=5, fig.height=4}
# Plot P(chose nearest) by distance, split by HF/LF nearest and PAYOFF structure
n_bins_payoff <- 9

# Filter to 0-22.5 range and bin by payoff structure
d_binned_chose_nearest_payoff <- d_chose_nearest_exp2 |>
  filter(distance <= 22.5) |>
  mutate(dist_bin = cut(distance, breaks = n_bins_payoff, include.lowest = TRUE)) |>
  group_by(nearestFreq, payoff_structure, dist_bin) |>
  tidyboot_mean(chose_nearest) |>
  left_join(
    d_chose_nearest_exp2 |>
      filter(distance <= 22.5) |>
      mutate(dist_bin = cut(distance, breaks = n_bins_payoff, include.lowest = TRUE)) |>
      group_by(nearestFreq, payoff_structure, dist_bin) |>
      summarise(dist_mean = mean(distance), .groups = "drop"),
    by = c("nearestFreq", "payoff_structure", "dist_bin")
  )

# Model predictions for payoff effect
pred_chose_nearest_payoff <- expand.grid(
  distance = seq(0, 22.5, length.out = 100),
  is_hf_trial_c = c(-0.5, 0.5),
  payoff_c = c(-0.5, 0.5),
  time_pressure_c = 0  # Average across time pressure conditions
) |>
  mutate(
    distance_c = (distance - dist_mean_cn) / dist_sd_cn,
    nearestFreq = ifelse(is_hf_trial_c > 0, "HF", "LF"),
    payoff_structure = ifelse(payoff_c < 0, "linear", "binary"),
    p_chose_nearest = predict(m_chose_nearest_exp2, newdata = pick(everything()), 
                               type = "response", re.form = NA)
  )

ggplot(d_binned_chose_nearest_payoff, aes(x = dist_mean, y = empirical_stat, 
                                    color = nearestFreq, shape = payoff_structure)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.5) +
  geom_line(data = pred_chose_nearest_payoff, 
            aes(x = distance, y = p_chose_nearest, linetype = payoff_structure),
            linewidth = 1) +
  scale_color_manual(values = c("HF" = "#E94F37", "LF" = "#2E86AB")) +
  scale_linetype_manual(values = c("linear" = "solid", "binary" = "dashed"),
                        labels = c("Linear", "Binary")) +
  scale_shape_manual(values = c("linear" = 16, "binary" = 17),
                     labels = c("Linear", "Binary")) +
  scale_x_continuous(limits = c(0, 22.5), breaks = seq(0, 22.5, by = 5)) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.25)) +
  labs(
    x = "Distance from Nearest Trained Angle (deg)",
    y = "P(Chose Nearest)",
    color = "Nearest Freq",
    shape = "Payoff Structure",
    linetype = "Payoff Structure",
    title = "Exp 2: P(Chose Nearest) by Distance and Target Frequency (Payoff)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "right",
    legend.background = element_rect(fill = "white", color = "gray80")
  )
```

# Figure: Payoff Effect

```{r fig-setup}
# Bin data for plotting
n_bins <- 9

# Common plot theme
theme_fig <- theme_minimal(base_size = 12) +
  theme(
    legend.position = c(0.75, 0.85),
    legend.background = element_rect(fill = "white", color = NA),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )
```

```{r fig-payoff, fig.width=5, fig.height=4}
# Panel A: Time Pressure effect
d_binned_tp_exp2 <- d_critical_exp2 |>
  mutate(dist_bin = cut(dist_to_lf, breaks = n_bins, include.lowest = TRUE)) |>
  group_by(time_pressure, dist_bin) |>
  tidyboot_mean(responded_lf) |>
  left_join(
    d_critical_exp2 |>
      mutate(dist_bin = cut(dist_to_lf, breaks = n_bins, include.lowest = TRUE)) |>
      group_by(time_pressure, dist_bin) |>
      summarise(dist_mean = mean(dist_to_lf), .groups = "drop"),
    by = c("time_pressure", "dist_bin")
  )

# Model predictions
# Need to scale dist_to_lf using the same scaling parameters from the data
dist_mean_exp2 <- mean(d_critical_exp2$dist_to_lf)
dist_sd_exp2 <- sd(d_critical_exp2$dist_to_lf)

pred_tp_exp2 <- expand.grid(
  dist_to_lf = seq(0, 45, length.out = 100),
  time_pressure_c = c(-0.5, 0.5),
  payoff_c = 0
) |>
  mutate(
    dist_to_lf_scaled = (dist_to_lf - dist_mean_exp2) / dist_sd_exp2,
    time_pressure = ifelse(time_pressure_c < 0, 5, 10),
    p_lf = predict(m_tp_exp2, newdata = pick(everything()), type = "response", re.form = NA)
  )

p_tp_exp2 <- ggplot(d_binned_tp_exp2, aes(x = dist_mean, y = empirical_stat, color = factor(time_pressure))) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 1) +
  geom_line(data = pred_tp_exp2, aes(x = dist_to_lf, y = p_lf), linewidth = 1.2) +
  geom_vline(xintercept = 22.5, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = colors_tp, labels = c("Strict (5s)", "Lenient (10s)")) +
  scale_x_continuous(limits = c(0, 45), breaks = seq(0, 45, by = 10)) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.25)) +
  labs(x = "Distance from LF (deg)", y = "P(LF)", color = "Time Pressure", title = "A") +
  theme_fig


p_tp_exp2
```

```{r}
# Panel B: Payoff Structure effect
d_binned_payoff_exp2 <- d_critical_exp2 |>
  mutate(dist_bin = cut(dist_to_lf, breaks = n_bins, include.lowest = TRUE)) |>
  group_by(payoff_structure, dist_bin) |>
  tidyboot_mean(responded_lf) |>
  left_join(
    d_critical_exp2 |>
      mutate(dist_bin = cut(dist_to_lf, breaks = n_bins, include.lowest = TRUE)) |>
      group_by(payoff_structure, dist_bin) |>
      summarise(dist_mean = mean(dist_to_lf), .groups = "drop"),
    by = c("payoff_structure", "dist_bin")
  )

# Model predictions
pred_payoff_exp2 <- expand.grid(
  dist_to_lf = seq(0, 45, length.out = 100),
  payoff_c = c(-0.5, 0.5),
  time_pressure_c = 0
) |>
  mutate(
    dist_to_lf_scaled = (dist_to_lf - dist_mean_exp2) / dist_sd_exp2,
    payoff_structure = ifelse(payoff_c < 0, "linear", "binary"),
    p_lf = predict(m_payoff_exp2, newdata = pick(everything()), type = "response", re.form = NA)
  )

p_payoff_exp2 <- ggplot(d_binned_payoff_exp2, aes(x = dist_mean, y = empirical_stat, color = payoff_structure)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 1) +
  geom_line(data = pred_payoff_exp2, aes(x = dist_to_lf, y = p_lf), linewidth = 1.2) +
  geom_vline(xintercept = 22.5, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = colors_payoff) +
  scale_x_continuous(limits = c(0, 45), breaks = seq(0, 45, by = 10)) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.25)) +
  labs(x = "Distance from LF (deg)", y = "P(LF)", color = "Payoff Structure", title = "B") +
  theme_fig
p_payoff_exp2
```

```{r fig-combined, fig.width=10, fig.height=4}
#| fig-cap: "P(LF) by distance from LF prototype"

p_combined_exp2 <- p_tp_exp2 + p_payoff_exp2
p_combined_exp2

ggsave(file.path(fig_path, "empirical_results_exp2.pdf"), p_combined_exp2,
       width = 10, height = 4, dpi = 300)
```

# Combined Model: Time Pressure × Payoff Structure

```{r model-combined}
# Full model with both two-way interactions
# Random effects: by-participant intercepts and slopes for distance (uncorrelated), by-word intercepts
m_combined_exp2 <- glmer(
  responded_lf ~ (time_pressure_c + payoff_c) * dist_to_lf_scaled + 
    (1 + dist_to_lf_scaled | subject_id) + (1 | lf_word),
  data = d_critical_exp2, family = binomial,
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

summary(m_combined_exp2)
```

```{r fig-combined-plot, fig.width=6, fig.height=5}
#| fig-cap: "Combined effects of time pressure and payoff structure"

# Bin data by both factors
d_binned_combined_exp2 <- d_critical_exp2 |>
  mutate(dist_bin = cut(dist_to_lf, breaks = n_bins, include.lowest = TRUE)) |>
  group_by(time_pressure, payoff_structure, dist_bin) |>
  tidyboot_mean(responded_lf) |>
  left_join(
    d_critical_exp2 |>
      mutate(dist_bin = cut(dist_to_lf, breaks = n_bins, include.lowest = TRUE)) |>
      group_by(time_pressure, payoff_structure, dist_bin) |>
      summarise(dist_mean = mean(dist_to_lf), .groups = "drop"),
    by = c("time_pressure", "payoff_structure", "dist_bin")
  )

# Model predictions for all condition combinations
pred_combined_exp2 <- expand.grid(
  dist_to_lf = seq(0, 45, length.out = 100),
  time_pressure_c = c(-0.5, 0.5),
  payoff_c = c(-0.5, 0.5)
) |>
  mutate(
    dist_to_lf_scaled = (dist_to_lf - dist_mean_exp2) / dist_sd_exp2,
    time_pressure = ifelse(time_pressure_c < 0, 5, 10),
    payoff_structure = ifelse(payoff_c < 0, "linear", "binary"),
    p_lf = predict(m_combined_exp2, newdata = pick(everything()), type = "response", re.form = NA)
  )

# Create plot with color = payoff, linetype = time pressure
p_combined_effects_exp2 <- ggplot() +
  geom_point(data = d_binned_combined_exp2, 
             aes(x = dist_mean, y = empirical_stat, 
                 color = payoff_structure, shape = factor(time_pressure)),
             size = 2.5, alpha = 0.7) +
  geom_line(data = pred_combined_exp2, 
            aes(x = dist_to_lf, y = p_lf, 
                color = payoff_structure, linetype = factor(time_pressure)),
            linewidth = 1) +
  geom_vline(xintercept = 22.5, linetype = "dotted", color = "gray50") +
  scale_x_continuous(limits = c(0, 45), breaks = seq(0, 45, by = 10)) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.25)) +
  labs(x = "Distance from LF (deg)", y = "P(LF)") +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "right",
    legend.background = element_rect(fill = "white", color = "gray80"),
    panel.grid.minor = element_blank()
  )

p_combined_effects_exp2

ggsave(file.path(fig_path, "combined_effects_exp2.pdf"), p_combined_effects_exp2,
       width = 7, height = 5, dpi = 300)
```

# Race Model Plots (separate document)

Race model fit plots (P(LF) vs distance, schematic, P(LF)/P(HF) vs time) are in **`experiment2_race_model.qmd`**.

**To generate them:**

1. From the repo root: `python models/fit_exp2.py` (writes CSVs to `models/`).
2. Render `experiment2_race_model.qmd` (e.g. Quarto render or "Run All" in that file).

No other Experiment 2 data or analysis is needed for the race model document.

# Response Time Analysis

```{r rt-by-condition-payoff-table}
# Mean RT (s) by incentive x frequency x time pressure for the payoff/parameter table
rt_payoff_table <- d_critical_exp2 |>
  mutate(rt_s = rt / 1000) |>
  group_by(frequency_ratio, payoff_structure, time_pressure) |>
  summarise(mean_rt = mean(rt_s), .groups = "drop") |>
  pivot_wider(
    names_from = time_pressure,
    values_from = mean_rt,
    names_prefix = "rt_"
  )

# Fitted T values (e.g. from fit_exp2 with sigma=0.20, delta=1 fixed); update if using different fit
T_payoff_table <- tibble(
  frequency_ratio = c("1:2", "1:2", "1:4", "1:4"),
  payoff_structure = c("linear", "binary", "linear", "binary"),
  T_5 = c(1.43, 1.63, 1.71, 1.67),
  T_10 = c(1.69, 1.76, 1.63, 1.87)
)

payoff_table <- T_payoff_table |>
  left_join(
    rt_payoff_table,
    by = c("frequency_ratio", "payoff_structure")
  ) |>
  rename(mean_rt_5 = rt_5, mean_rt_10 = rt_10)

# LaTeX table: FR, Payoff, T(5s), T(10s), Mean RT (5s), Mean RT (10s)
payoff_table_latex <- payoff_table |>
  mutate(
    FR = frequency_ratio,
    Payoff = str_to_title(payoff_structure),
    across(c(T_5, T_10, mean_rt_5, mean_rt_10), \(x) sprintf("%.2f", x))
  ) |>
  select(FR, Payoff, T_5, T_10, mean_rt_5, mean_rt_10)
```

**Payoff table with fitted $T$ and mean human RT (s)** — paste into manuscript:

```{r payoff-table-latex, echo=FALSE}
cat("\\\\begin{table}[t]\n")
cat("\\\\centering\n")
cat("\\\\small\n")
cat("\\\\begin{tabular}{llcccc}\n")
cat("\\\\hline\n")
cat("FR & Payoff & $T$ (5s) & $T$ (10s) & Mean RT (5s) & Mean RT (10s) \\\\\n")  # \\ in output
cat("\\\\hline\n")
for (i in 1:nrow(payoff_table_latex)) {
  cat(paste(payoff_table_latex$FR[i], payoff_table_latex$Payoff[i],
            payoff_table_latex$T_5[i], payoff_table_latex$T_10[i],
            payoff_table_latex$mean_rt_5[i], payoff_table_latex$mean_rt_10[i],
            sep = " & "), "\\\\\\\\\n")  # 4 backslashes in R -> \\ in output for LaTeX line break
}
cat("\\\\hline\n")
cat("\\\\end{tabular}\n")
cat("\\\\caption{Effective decision time parameter $T$ fits by condition ($\\\\sigma = 0.20$ and $\\\\delta=1$ fixed) and mean human RT (s) by condition.}\n")
cat("\\\\label{tab:exp2params}\n")
cat("\\\\end{table}\n")
```

```{r rt-summary}
d_critical_exp2 |>
  group_by(time_pressure, responseFreq) |>
  summarise(M = mean(rt), SD = sd(rt), n = n(), .groups = "drop") |>
  knitr::kable(digits = 0, col.names = c("Time Pressure", "Response", "M RT", "SD", "N"))
```

```{r rt-summary-payoff}
d_critical_exp2 |>
  group_by(payoff_structure, responseFreq) |>
  summarise(M = mean(rt), SD = sd(rt), n = n(), .groups = "drop") |>
  knitr::kable(digits = 0, col.names = c("Reward Structure", "Response", "M RT", "SD", "N"))
```

```{r rt-model}
m_rt_exp2 <- lmer(
  log(rt) ~ responded_lf * dist_to_lf_scaled * payoff_c + 
    (1 + dist_to_lf_scaled | subject_id) + (1 | lf_word),
  data = d_critical_exp2,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)
summary(m_rt_exp2)
```

# Summary

```{r summary}
cat("=== EXPERIMENT 2 SUMMARY ===\n\n")
cat("N =", n_participants, "participants\n")
cat("N =", nrow(d_critical_exp2), "critical trials\n")
cat("N =", n_distinct(d_critical_exp2$lf_word), "unique LF words\n\n")
cat("Timeout rates:\n")
cat(sprintf("  Strict (5s): %.1f%%\n", timeout_by_tp_exp2$pct_timeout[1]))
cat(sprintf("  Lenient (10s): %.1f%%\n", timeout_by_tp_exp2$pct_timeout[2]))
cat("\nKey interactions (scaled distance):\n")
cat(sprintf("  Time pressure x distance: B = %.3f, z = %.2f, p %s\n", 
            tp_coef_exp2["Estimate"], tp_coef_exp2["z value"],
            ifelse(tp_coef_exp2["Pr(>|z|)"] < .001, "< .001", sprintf("= %.3f", tp_coef_exp2["Pr(>|z|)"]))))
cat(sprintf("  Payoff Condition x distance: B = %.3f, z = %.2f, p %s\n", 
            payoff_coef_exp2["Estimate"], payoff_coef_exp2["z value"],
            ifelse(payoff_coef_exp2["Pr(>|z|)"] < .001, "< .001", sprintf("= %.3f", payoff_coef_exp2["Pr(>|z|)"]))))
```

# Supplementary: Response Proportions by Target Frequency

This analysis examines the proportion of HF, LF, and Other responses as a function of which frequency word is nearest to the target angle. "Other" responses are valid trained words that are neither the nearest nor next-nearest option. Restricted to ambiguous trials (>11° from nearest trained angle) to focus on cases where the target is clearly between two options.

```{r response-proportions-exp2, fig.width=8, fig.height=6}
# Add response category: nearest, next-nearest, or other valid word
d_exp2_with_category <- d_exp2 |>
  filter(between_freqs, ambig_trial, timed_out == 0, !is.na(responseFreq)) |>
  mutate(
    response_category = case_when(
      response_text == nearestLabel ~ nearestFreq,
      response_text == nextNearestLabel ~ nextNearestFreq,
      TRUE ~ "Other"
    )
  )

# Compute response proportions on ambiguous between-frequency trials (aggregate)
d_props_exp2 <- d_exp2_with_category |>
  group_by(nearestFreq, payoff_structure, time_pressure) |>
  summarise(
    n = n(),
    n_lf = sum(response_category == "LF"),
    n_hf = sum(response_category == "HF"),
    n_other = sum(response_category == "Other"),
    prop_lf = n_lf / n,
    prop_hf = n_hf / n,
    prop_other = n_other / n,
    .groups = "drop"
  )

# By-participant proportions for individual points
d_props_by_subj_exp2 <- d_exp2_with_category |>
  group_by(subject_id, nearestFreq, payoff_structure, time_pressure) |>
  summarise(
    n = n(),
    prop_lf = sum(response_category == "LF") / n,
    prop_hf = sum(response_category == "HF") / n,
    prop_other = sum(response_category == "Other") / n,
    .groups = "drop"
  ) |>
  pivot_longer(cols = c(prop_lf, prop_hf, prop_other), 
               names_to = "response_type", 
               values_to = "proportion") |>
  mutate(
    response_type = case_when(
      response_type == "prop_lf" ~ "LF",
      response_type == "prop_hf" ~ "HF",
      response_type == "prop_other" ~ "Other"
    ),
    response_type = factor(response_type, levels = c("HF", "LF", "Other")),
    time_pressure = factor(time_pressure, labels = c("5s (Strict)", "10s (Lenient)")),
    payoff_structure = factor(payoff_structure, 
                              levels = c("linear", "binary"),
                              labels = c("Linear Payoff", "Binary Payoff"))
  )

# Reshape aggregate for plotting
d_props_long_exp2 <- d_props_exp2 |>
  pivot_longer(cols = c(prop_lf, prop_hf, prop_other), 
               names_to = "response_type", 
               values_to = "proportion") |>
  mutate(
    response_type = case_when(
      response_type == "prop_lf" ~ "LF",
      response_type == "prop_hf" ~ "HF",
      response_type == "prop_other" ~ "Other"
    ),
    response_type = factor(response_type, levels = c("HF", "LF", "Other")),
    time_pressure = factor(time_pressure, labels = c("5s (Strict)", "10s (Lenient)")),
    payoff_structure = factor(payoff_structure, 
                              levels = c("linear", "binary"),
                              labels = c("Linear Payoff", "Binary Payoff"))
  )

# Plot: Response proportions by nearest frequency with participant points
ggplot(d_props_long_exp2, aes(x = nearestFreq, y = proportion, fill = response_type)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.6) +
  geom_point(data = d_props_by_subj_exp2,
             aes(x = nearestFreq, y = proportion, color = response_type),
             position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.9),
             alpha = 0.4, size = 1.5, show.legend = FALSE) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray50") +
  facet_grid(time_pressure ~ payoff_structure) +
  scale_fill_manual(values = c("HF" = "#E94F37", "LF" = "#2E86AB", "Other" = "#7D7D7D")) +
  scale_color_manual(values = c("HF" = "#B33D2B", "LF" = "#1F5F7A", "Other" = "#4D4D4D")) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.25)) +
  labs(
    x = "Nearest Trained Angle Frequency",
    y = "Proportion of Responses",
    fill = "Response",
    title = "Exp 2 Response Proportions",
    subtitle = "Bars = group means; Points = individual participants; Other = valid word but not nearest/next-nearest"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold")
  )
```

```{r response-table-exp2}
# Summary table
d_props_exp2 |>
  mutate(
    time_pressure = factor(time_pressure, labels = c("5s", "10s"))
  ) |>
  knitr::kable(
    digits = 2,
    col.names = c("Nearest Freq", "Payoff Structure", "Time Pressure", "N Trials", 
                  "N LF", "N HF", "N Other", "P(LF)", "P(HF)", "P(Other)"),
    caption = "Response proportions on ambiguous between-frequency trials"
  )
```

# Exploratory: Learning Quality Checks

This section examines accuracy on exact-angle trials (timed production at trained angles) and untimed final recall trials to assess overall learning quality.

```{r load-accuracy-data-exp2}
# Need trial_type to identify exact_angle trials, and response to identify final-recall
cols_accuracy <- c("subject_id", "payoff_structure", "time_pressure", "trial_type",
                   "is_correct", "response", "target", "angle", "targetFreq")

# Load accuracy data from both sources (same split as main data)
accuracy_data_binary <- map_dfr(binary_files, \(f) {
  read_csv(f, show_col_types = FALSE) |>
    select(any_of(cols_accuracy)) |>
    mutate(across(everything(), as.character))
}) |>
  filter(payoff_structure == "binary")

accuracy_data_linear <- map_dfr(linear_files, \(f) {
  read_csv(f, show_col_types = FALSE) |>
    select(any_of(cols_accuracy)) |>
    mutate(across(everything(), as.character))
}) |>
  filter(payoff_structure == "linear")

accuracy_data_exp2 <- bind_rows(accuracy_data_binary, accuracy_data_linear)

# Extract exact_angle trials
exact_angle_exp2 <- accuracy_data_exp2 |>
  filter(trial_type == "exact_angle") |>
  transmute(
    subject_id,
    is_correct = as.numeric(is_correct),
    target,
    angle = as.numeric(angle),
    targetFreq
  ) |>
  left_join(participants, by = "subject_id")

# Extract final recall trials (untimed)
final_recall_exp2 <- accuracy_data_exp2 |>
  filter(str_detect(response, "final-recall-response")) |>
  transmute(
    subject_id,
    is_correct = as.numeric(is_correct),
    target,
    angle = as.numeric(angle),
    targetFreq
  ) |>
  left_join(participants, by = "subject_id")

cat("Exact angle trials:", nrow(exact_angle_exp2), "\n")
cat("Final recall trials:", nrow(final_recall_exp2), "\n")
```



```{r exact-angle-by-targetfreq-exp2, fig.width=8, fig.height=6}
# Exact angle trials: accuracy by target frequency (HF vs LF)
exact_acc_by_freq_exp2 <- exact_angle_exp2 |>
  group_by(targetFreq, payoff_structure, time_pressure) |>
  summarise(
    n_trials = n(),
    accuracy = mean(is_correct, na.rm = TRUE),
    .groups = "drop"
  )

# By-participant accuracy for individual points
exact_acc_by_freq_subj_exp2 <- exact_angle_exp2 |>
  group_by(subject_id, targetFreq, payoff_structure, time_pressure) |>
  summarise(
    accuracy = mean(is_correct, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(exact_acc_by_freq_exp2, aes(x = targetFreq, y = accuracy, fill = targetFreq)) +
  geom_col(alpha = 0.6) +
  geom_point(data = exact_acc_by_freq_subj_exp2,
             aes(x = targetFreq, y = accuracy, color = targetFreq),
             position = position_jitter(width = 0.2),
             alpha = 0.4, size = 1.5, show.legend = FALSE) +
  facet_grid(time_pressure ~ payoff_structure, 
             labeller = labeller(
               time_pressure = c("5" = "5s (Strict)", "10" = "10s (Lenient)"),
               payoff_structure = c("linear" = "Linear Payoff", "binary" = "Binary Payoff")
             )) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = c("HF" = "#E94F37", "LF" = "#2E86AB")) +
  scale_color_manual(values = c("HF" = "#B33D2B", "LF" = "#1F5F7A")) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.25)) +
  labs(
    x = "Target Frequency",
    y = "Accuracy",
    fill = "Target Freq",
    title = "Exp 2: Exact Angle Trial Accuracy by Target Frequency",
    subtitle = "Timed production at trained angles; Bars = group means; Points = individual participants"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold")
  )
```

```{r final-recall-by-targetfreq-exp2, fig.width=8, fig.height=6}
# Final recall trials: accuracy by target frequency (HF vs LF)
recall_acc_by_freq_exp2 <- final_recall_exp2 |>
  group_by(targetFreq, payoff_structure, time_pressure) |>
  summarise(
    n_trials = n(),
    accuracy = mean(is_correct, na.rm = TRUE),
    .groups = "drop"
  )

# By-participant accuracy for individual points
recall_acc_by_freq_subj_exp2 <- final_recall_exp2 |>
  group_by(subject_id, targetFreq, payoff_structure, time_pressure) |>
  summarise(
    accuracy = mean(is_correct, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(recall_acc_by_freq_exp2, aes(x = targetFreq, y = accuracy, fill = targetFreq)) +
  geom_col(alpha = 0.6) +
  geom_point(data = recall_acc_by_freq_subj_exp2,
             aes(x = targetFreq, y = accuracy, color = targetFreq),
             position = position_jitter(width = 0.2),
             alpha = 0.4, size = 1.5, show.legend = FALSE) +
  facet_grid(time_pressure ~ payoff_structure, 
             labeller = labeller(
               time_pressure = c("5" = "5s (Strict)", "10" = "10s (Lenient)"),
               payoff_structure = c("linear" = "Linear Payoff", "binary" = "Binary Payoff")
             )) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = c("HF" = "#E94F37", "LF" = "#2E86AB")) +
  scale_color_manual(values = c("HF" = "#B33D2B", "LF" = "#1F5F7A")) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.25)) +
  labs(
    x = "Target Frequency",
    y = "Accuracy",
    fill = "Target Freq",
    title = "Exp 2: Final Recall Accuracy by Target Frequency",
    subtitle = "Untimed production at trained angles; Bars = group means; Points = individual participants"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold")
  )
```

```{r accuracy-summary-exp2}
cat("\n=== ACCURACY SUMMARY (Exp 2) ===\n")
cat(sprintf("Exact Angle (Timed): %.1f%%\n", 100 * exact_acc_overall_exp2$accuracy))
cat(sprintf("Final Recall (Untimed): %.1f%%\n", 100 * recall_acc_overall_exp2$accuracy))
```
