---
title: "Experiment 2: Benefit Pressure"
author: "Elves Project Team"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 8
    fig-height: 5
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(lme4)
library(lmerTest)
library(here)
library(scales)

# Color palette for payoff conditions
# TODO: adjust colors based on actual conditions
colors_payoff <- c("low" = "#E94F37", "high" = "#2E86AB")

# Output paths
fig_path <- here("latex/figures")
```

# Overview

Experiment 2 tests whether speakers adjust their retrieval strategies when precision is more valuable. The key manipulation is the **payoff for precise responses**.

**Key prediction**: Higher payoff for precision → steeper boundaries (more LF at prototype, similar HF at boundaries).

# Data Loading

```{r load-data}
# TODO: Update paths once Exp2 data is collected
# participants <- read_csv(here("data/exp2/processed_data/participants.csv"),
#                          show_col_types = FALSE)
# production <- read_csv(here("data/exp2/processed_data/production.csv"),
#                        show_col_types = FALSE)

# Placeholder message
cat("Awaiting Experiment 2 data...\n")
cat("Expected data path: data/exp2/processed_data/\n")
```

# Sample Descriptives

```{r sample-descriptives}
# TODO: Fill in once data available
# n_participants <- n_distinct(participants$subject_id)
#
# condition_counts <- participants |>
#   count(payoff_condition) |>
#   pivot_wider(names_from = payoff_condition, values_from = n)
#
# cat("Total participants:", n_participants, "\n\n")
# cat("Condition breakdown:\n")
# knitr::kable(condition_counts)
```

# Critical Trial Selection

```{r filter-critical}
# TODO: Adapt from Exp1 based on Exp2 design
#
# d_critical <- d |>
#   filter(
#     between_freqs == TRUE,
#     timed_out == 0,
#     !is.na(responseFreq),
#     rt > 200
#   ) |>
#   mutate(
#     responded_lf = as.numeric(responseFreq == "LF"),
#     abs_distance = abs(distance),
#     # TODO: define payoff condition coding
#     payoff_c = ifelse(payoff_condition == "low", -0.5, 0.5)
#   )
```

# Figure: Payoff Manipulation Effect

```{r fig-payoff, fig.width=5, fig.height=4}
#| label: fig-payoff
#| fig-cap: "P(LF) by distance from LF trained angle, split by payoff condition"

# TODO: Create figure similar to Exp1 Panel A
#
# n_bins <- 9
# d_binned <- d_critical |>
#   mutate(dist_bin = cut(abs_distance, breaks = n_bins, include.lowest = TRUE)) |>
#   group_by(payoff_condition, dist_bin) |>
#   summarise(
#     n = n(),
#     p_lf = mean(responded_lf),
#     se = sqrt(p_lf * (1 - p_lf) / n),
#     dist_mean = mean(abs_distance),
#     .groups = "drop"
#   )
#
# # Fit model for smooth curves
# m_payoff <- glmer(
#   responded_lf ~ abs_distance * payoff_c + (1 | subject_id),
#   data = d_critical,
#   family = binomial(link = "logit")
# )
#
# # Plot
# p_payoff <- ggplot(d_binned, aes(x = dist_mean, y = p_lf,
#                                   color = payoff_condition)) +
#   geom_point(size = 3) +
#   geom_errorbar(aes(ymin = p_lf - se, ymax = p_lf + se), width = 0.5) +
#   geom_vline(xintercept = 22.5, linetype = "dashed", color = "gray50") +
#   scale_color_manual(values = colors_payoff) +
#   scale_x_continuous(limits = c(0, 23)) +
#   scale_y_continuous(limits = c(0, 1)) +
#   labs(
#     x = "Distance from LF (°)",
#     y = "P(LF)",
#     color = "Payoff"
#   ) +
#   theme_minimal(base_size = 12) +
#   theme(legend.position = c(0.75, 0.85))
#
# p_payoff

cat("Figure placeholder: awaiting data\n")
```

# Statistical Models

## Model 1: Main Effect of Payoff

```{r model-payoff}
# TODO: Run models once data available
#
# m1 <- glmer(
#   responded_lf ~ payoff_c + distance_c + (1 | subject_id),
#   data = d_critical,
#   family = binomial(link = "logit")
# )
#
# summary(m1)
```
## Model 2: Payoff × Distance Interaction

**Key test**: Does payoff condition interact with distance from LF prototype?

```{r model-interaction}
# TODO: Run interaction model
#
# m2 <- glmer(
#   responded_lf ~ payoff_c * distance_c + (1 | subject_id),
#   data = d_critical,
#   family = binomial(link = "logit")
# )
#
# summary(m2)
#
# # Model comparison
# anova(m1, m2)
```

**Expected result**: Significant payoff × distance interaction, with higher payoff producing steeper slope (more differentiation between prototype and boundary).

# Response Time Analysis

```{r rt-analysis}
# TODO: RT analysis
#
# rt_summary <- d_critical |>
#   group_by(payoff_condition) |>
#   summarise(
#     M_rt = mean(rt),
#     SD_rt = sd(rt),
#     .groups = "drop"
#   )
#
# knitr::kable(rt_summary)
```

# Comparison with Experiment 1

```{r exp-comparison}
# TODO: Load Exp1 data and compare effect sizes
#
# Effect size comparison:
# - Exp1 time pressure effect
# - Exp1 frequency ratio effect
# - Exp2 payoff effect
```

# Summary Statistics for Paper

```{r summary-for-paper}
cat("=== SUMMARY FOR MANUSCRIPT ===\n\n")
cat("Awaiting Experiment 2 data collection...\n")

# TODO: Fill in once data available
# cat("Participants:\n")
# cat(sprintf("  N = %d\n", n_participants))
#
# cat("\nPayoff × distance interaction:\n")
# payoff_int <- summary(m2)$coefficients["payoff_c:distance_c", ]
# cat(sprintf("  β = %.3f, z = %.2f, p = %.3f\n",
#             payoff_int["Estimate"], payoff_int["z value"], payoff_int["Pr(>|z|)"]))
```

# Session Info

```{r session-info}
sessionInfo()
```
